---
title: "Parsimonious Approach to Gerrymandering Analysis"
author: "Joseph Holler"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  markdown:
    wrap: sentence
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../docs") })
nocite: '@*'
bibliography: "../../software.bib"
---

# Abstract

This is a study of gerrymandering in Alabama.
We will test three methods of shape-based compactness scores, assess representativeness of districts based on prior presidential elections and race. 
We will then extend prior studies by calculating representativeness of the convex hull of district polygons.

## Study Metadata

- `Key words`: Alabama, gerrymandering, compactness, convex hull, political representation
- `Subject`: Social and Behavioral Sciences: Geography: Geographic Information Sciences
- `Date created`: 2025-02-17
- `Date modified`: 2025-02-17
- `Spatial Coverage`: Alabama OSM:[161950](https://www.openstreetmap.org/relation/161950)
- `Spatial Resolution`: Census Block Groups
- `Spatial Reference System`: EPSG:4269 NAD 1983 Geographic Coordinate System
- `Temporal Coverage`: 2020-2024 population and voting data
- `Temporal Resolution`: Decennial census

# Study design

This is an original study based on literature on gerrymandering metrics.

It is an exploratory study to evaluate usefulness of a new gerrymandering metric based on the convex hull of a congressional district and the representativeness inside the convex hull compared to the congressional district.

# Materials and procedure

## Computational environment

I plan on using package ... for ... 

```{r environment-setup, include = FALSE}
# record all the packages you are using here
# this includes any calls to library(), require(),
# and double colons such as here::i_am()
packages <- c("tidyverse",
              "here",
              "sf",
              "lwgeom",
              "tmap",
              "tidycensus",
              "knitr", 
              "htmltools",
              "markdown",
              "corrplot")

# force all conflicts to become errors
# if you load dplyr and use filter(), R has to guess whether you mean dplyr::filter() or stats::filter()
# the conflicted package forces you to be explicit about this
# disable at your own peril
# https://conflicted.r-lib.org/
require(conflicted)

# load and install required packages
# https://groundhogr.com/
if (!require(groundhog)) {
  install.packages("groundhog")
  require(groundhog)
}

# this date will be used to determine the versions of R and your packages
# it is best practice to keep R and its packages up to date
groundhog.day <- "2026-06-30"


# this replaces any library() or require() calls
groundhog.library(packages, groundhog.day)
# you may need to install a correct version of R
# you may need to respond OK in the console to permit groundhog to install packages
# you may need to restart R and rerun this code to load installed packages
# In RStudio, restart r with Session -> Restart Session
```


```{r}
# record the R processing environment
# alternatively, use devtools::session_info() for better results
writeLines(
  capture.output(sessionInfo()),
  here("procedure", "environment", paste0("r-environment-", Sys.Date(), ".txt"))
)

# save package citations
knitr::write_bib(c(packages, "base"), file = here("software.bib"))

# set up default knitr parameters
# https://yihui.org/knitr/options/
knitr::opts_chunk$set(
  echo = TRUE, # Run code, show outputs (don't show code)
  fig.retina = 4,
  fig.width = 8,
  fig.path = paste0(here("results", "figures"), "/")
)
```

## Data and variables

### Decennial Census

We acquire decennial census data in block groups using the `tidycensus` package.
First, query metadata for the `pl` public law data series.
Then, load Alabama 2021 districts, 2024 precincts, and 2020 block groups.

```{r}
result_set <- "LA2024v2"
# districts_file <- here("data", "derived", "public", "districts_LA_2024_clipped.gpkg")
districts_file <- here("data", "raw", "public", "Act_2_1st_ES_(2024)_-_Congress.shp")
blockgroup_file <- here("data", "raw", "public", "LA_block_groups.gpkg")
# precincts_file <- here("data", "derived", "public", "la_2024_gen_all_prec", "la_2024_gen_all_prec.shp")
precincts_file <- here("data", "derived", "public", "LA_precincts.gpkg")


districts <- st_read(districts_file)
precincts <- st_read(precincts_file, layer = "LA_precincts_clipped")

census_metadata_file <- here("data", "metadata", "census2020pl_vars.csv")
if(file.exists(census_metadata_file)){
  census2020pl_vars <- read.csv(census_metadata_file)
} else {
  census2020pl_vars <- load_variables(2020, "pl")
  write.csv(census2020pl_vars, here("data", "metadata", "census2020pl_vars.csv"))
}

# if the data is already downloaded, just load it
# otherwise, query from the census and save
if(file.exists(blockgroup_file)){
  blockgroups <- st_read(blockgroup_file)
} else {
  blockgroups <- get_decennial(geography = "block group",
                               sumfile = "pl",
                               table = "P3",
                               year = 2020,
                               state = "Louisiana",
                               output = "wide",
                               geometry = TRUE,
                               keep_geo_vars = TRUE)
  st_write(blockgroups, blockgroup_file)
}
```

### Districts

The congressional districts enacted in 2021 were used in the 2022 mid-term elections, and then deemed unconstitutional gerrymanders by the Supreme Court.


Map the districts

```{r}
plot_district_map <- function(df, col_var = "red") {
  tmap_mode(mode = "plot")

  df |>
    tm_shape() +
    tm_polygons(fill_alpha = 0,
                col = col_var,
                col_alpha = 0.5) +
    tm_text(text = "DISTRICT",
            col = "black",
            #  bgcol = "white",
            bgcol_alpha = 0.7,
            options = opt_tm_text(on_surface = TRUE,
                                  halo = TRUE,
                                  halo.col = "white"))
}

map_districts <- plot_district_map(districts, "red")
map_districts
```


The issue in the 2023 court cases on Alabama's gerrymandering was a racial gerrymander discriminating against people identifying as Black or African American.
Therefore, we will analyze people of voting age (18 or older) identifying as Black and or African as one race in any combination with other races.
This data is found in table `P3`.

Query the public law data series table P3 on "race for the population 18 years and over".


## Prior observations  

We have previously investigated the compactness scores of Alabama's congressional districts as well as the percentage of Biden voters from the 2020 elections and the percentage of the population 18 years or older that is not Hispanic and is Black or African American.

We have never calculated the minimum bounding circle or convex hulls of Alabama's congressional districts.

## Bias and threats to validity

This study is explicitly an investigation to the modifiable areal unit problem.
Aspects of the study are extremely sensitive to the combination of edge effects and scale, whereby complex borders formed by natural features, e.g. coastlines or rivers, vary greatly in perimeter depending on the scale of analysis. 
We hope that in part, this study establishes a method that is more robust (less sensitive) to the threats to validity caused by scale and edge effects in studies of gerrymandering and district shapes.

## Data transformations

### Districts 2021

Transform Districts into NAD 1983 coordinate system, and calculate the percentage of population identifying as Black.

```{r}
districts <- districts |> st_transform(4269) 
st_geometry(districts) <- "geom"
# |> mutate(pctMinority = round(BLACK / POPULATION * 100, 1))
```


### Block groups census data

Census data needs to be transformed from the WGS 1984 geographic coordinate system to the NAD 1983 geographic coordinate system.

```{r}
blockgroups <- st_transform(blockgroups, 4269)
```

Find the total of people identifying as Black or African American as one race or any combination of multiple races.
First, make a list of all the variables inclusive of people identifying as Black or African American.  
**NOTE:** If we were to analyze multiple minority racial or ethnic groups, it may be wiser to focus on the `P4` table crosstabulating race and ethnicity and to focus on individuals identifying as only one race. This is to avoid double-counting people and simplify calculations.

```{r}
minority_cat <- "Black"

minority_vars <- census2020pl_vars |> 
  dplyr::filter(str_detect(name, "P3"),
                str_detect(label, minority_cat)) |> 
  select(-concept)

minority_vars |> kable()
```

Next, calculate new columns.
`Black` is a sum of all 32 columns shown above, in which any of the racial categories by which someone identifies is Black or African American.  
`Total` is a copy of the population 18 years or over, variable `P3_001N`.  
`pctMinority` is calculated as `Black / Total * 100`  
`CheckPct` is calculated as the percentage of the population 18 years or older that is either white of one race only (`P3_003N`) *or* Black or African American as calculated above. In Alabama and Louisiana, we can expect that this will be close to 100% for most block groups, and should never exceed 100%.

```{r}
blockgroups_calc <- blockgroups |> dplyr::filter(!st_is_empty(geom))
#st_geometry(blockgroups_calc) <- "geom" 
blockgroups_calc <- blockgroups_calc |> 
  rowwise() |> 
  mutate(minority = sum(c_across(all_of(minority_vars$name)))) |> 
  ungroup() |> 
  mutate(sarea = st_area(geom),
         Total = P3_001N,
         pctMinority = minority / Total * 100,
         CheckPct = (minority + P3_003N) / Total * 100
         ) |> 
  select(GEOID, sarea, minority, Total, pctMinority, CheckPct)
```

Save the results as `blockgroups_calc.gpkg`

```{r}
st_write(blockgroups_calc, 
         here("data", "derived", "public", "blockgroups_LA.gpkg"),
         append=FALSE)
```

Map the percentage of the population 18 or over that is a minority.

```{r message=FALSE}
tmap_mode(mode = "plot")
map_minority_race <- tm_shape(blockgroups_calc) + 
  tm_polygons(
    fill = "pctMinority",
    col_alpha = 0.2,
    lwd = 0.1,
    col = "grey90"
  )

map_minority_race + map_districts
```
Make an interactive map of the 2021 districts over the black population.

```{r warning=FALSE}
tmap_mode(mode = "view")
map_minority_race +
  map_districts
```

## Analysis

### District composition

- `race_fragments` are the intersection between Block Groups and Districts.
- `district_estimates` are tabular results by district
- `districts_join` are tabular results joined to the district polygons
- Minority (African American or Black) voting-age population is estimated with areal interpolation as `awMinor` in `race_fragments` and `bgMinority` in `districts_join`
- Total voting-age population is estimated with areal interpolation as `awTot` in `race_fragments` and `bgTotal` in `districts_join`

In the areal interpolation below, we take several steps to ensure that the 'fragments' are valid single-part polygons:

- `st_intersection` passes the option `dimensions="polygon"` to `s2_options`
- `sf` optimizes the *first* dataset in a spatial query or overlay with a spatial index, so the more complex race or voting data is first.
- `st_make_valid()` ensures valid geometries 
- `filter(!st_is_empty(geom))` removes results resulting in empty geometry collections
- `st_cast("MULTIPOLYGON") |> st_cast("POLYGON")` converts results to single-part polygons. It seems strange to convert to multipolygons first, but this is required in R `sf` to avoid losing parts of multi-part features when there are mixed geometry types. 

Areal interpolation / decomposition into fragments:

```{r warning=FALSE}
race_fragments <- st_intersection(blockgroups_calc, districts, dimensions="polygon") |> 
  st_make_valid() |> 
  dplyr::filter(!st_is_empty(geom)) |> 
  st_cast("MULTIPOLYGON") |> st_cast("POLYGON") |> 
  mutate(
    awTot = Total * as.numeric(st_area(geom) / sarea),
    awMinor = minority * as.numeric(st_area(geom) / sarea),
    frag_id = row_number()
  ) |> 
  select(DISTRICT, frag_id, sarea, Total = awTot, minority = awMinor)

```

Aggregation into districts:

```{r warning=FALSE}
district_estimates <- race_fragments |> 
  st_drop_geometry() |> 
  group_by(DISTRICT) |> 
  summarize(bgTotal = sum(Total),
            bgMinority = sum(minority))

districts_join <- districts |> 
  left_join(district_estimates, by = "DISTRICT") |> 
  mutate(pctMinoritybg = round(bgMinority / bgTotal * 100, 1))
```

Report results.
We find very similar percentages of Black or African American people using 
areal interpolation with block group data.

```{r}
districts_join |> st_drop_geometry() |> kable()
```

### Convex Hulls of Districts and their Composition

- `d_ch`: convex hulls of each district
- `d_ch_frag`: fragments from overlay of convex hulls with block groups
- `d_ch_est`: district-level summary table of convex hull composition
- Minority (African American or Black) voting-age population is estimated with areal interpolation as `awMinor` in `d_ch_frag` and `chMinority` in `d_ch_est`
- Total voting-age population is estimated with areal interpolation as `awTot` in `d_ch_frag` and `chTotal` in `d_ch_est`

NOTE: Filter fragments for polygons, not just valid geometries!

```{r convex_hull_race warning=FALSE}
d_ch <- districts |> 
  select(chDISTRICT = DISTRICT) |> 
  st_convex_hull() |> 
  mutate(chArea = st_area(geom))
  
d_ch_frag <- race_fragments |> 
  st_intersection(d_ch, dimensions="polygon") |> 
  st_make_valid() |> 
  dplyr::filter(!st_is_empty(geom)) |> 
  st_cast("MULTIPOLYGON") |> st_cast("POLYGON") |> 
  mutate(
    awTot = Total * as.numeric(st_area(geom) / sarea),
    awMinor = minority * as.numeric(st_area(geom) / sarea),
  )

d_ch_est <- d_ch_frag |> 
  st_drop_geometry() |> 
  group_by(chDISTRICT, chArea) |> 
  summarize(chTotal = sum(awTot, na.rm = TRUE),
            chMinority = sum(awMinor, na.rm = TRUE))
```

### District Cdiff scores

Join convex hull estimates to Districts with block group estimates.

- `chDiff` is the District composition minus the Convex Hull composition. A positive number indicates over-representation (packing?) in the district compared to the convex hull of the district, while a negative number represents an under-representation (dilution? cracking?) in the district compared to the convex hull of the district.
- `ChAbsDiff` is the absolute value of chDiff to measure total change in composition of the district compared to the convex hull of the district.

```{r}
districts_results <- districts_join |> 
  left_join(d_ch_est, by = c("DISTRICT" = "chDISTRICT")) |> 
  mutate(chPct = round(chMinority / chTotal * 100, 1),
         chDiff = pctMinoritybg - chPct,
         chAbsDiff = abs(chDiff)) 
```

### District Compactness scores

- `compact_shp` is the Posby-Popper metric based on the district area (`darea`) and perimeter (`dperim`)
- `compact_hull` is based on the district area (`darea`) and the area of the convex hull (`chArea`, calculated above)

```{r}
districts_results <- districts_results |> 
  mutate(
    darea = st_area(geom),
    dperim = st_length(st_cast(geom, "MULTILINESTRING")),
    compact_shp = round( as.numeric((4 * pi * darea) / dperim^2), 2),
    compact_hull = round( as.numeric(darea / chArea), 2),
  )
```

### Local Gerrymandering Scores

- Join District-level `chDiff` data to the `ch_frags` fragments from overlaying convex hulls with census geometries.
- `minorRatio` is the local minority ratio of the fragment
- `gerry_type` is the type of gerrymandering for the given fragment: `packing` when the fragment is inside of its district (`DISTRICT == chDISTRICT`) and `diluting` when he fragment is outside of its district (`DISTRICT != chDISTRICT`)
- `chRatio` is the minority ratio of the relevant district or convex hull, which is the chDiff of the district for packing and the chDiff of the convex hull (`ch_hull_Diff`) for diluting
- `gerry` represents total wasted minority votes due to gerrymandering, estimated in the packing form as the proportion of voting-age minorities in the fragment in excess of the proportion of voting-age minorities in the convex hull, and in the dilution form as the proportion of voting-age minorities in the fragment in excess of the proportion of voting-age minorities in the district.

**NOTE:** Currently there is no check for overlapping fragments. That means that the same location may contribute to packing in one distict *and* dilution/cracking in one or more districts. In a sense, wasted votes are potentially counted more than once. Does this make sense? It probably does, because wasting those votes is having an impact on elections in more than one district simultaneously.

**NOTE:** We also have not ensured that the total local wasted votes accounts for exactly the same shift in composition observed between the whole district and its convex hull. In order to do that, we would have to estimate the total wasted votes at the district level and then allocate them to the local areas in proportion to the local estimate of wasted local votes we have here. The total of local wasted votes would likely decrease while the spatial pattern would remain the same.

**IDEA:** One could add a flag/code for gerrymandering that crosses the 50% threshold for voting outcomes.

```{r wasted-votes}
districts_results_table <- districts_results |> st_drop_geometry() |> 
  select(DISTRICT, chDiff)

# join cDiff from district back to the fragment
ch_frags <- d_ch_frag |> 
  left_join(districts_results_table, by="DISTRICT")

# will need the cDiff of the convex hull polygon as well: ch_hull_Diff
districts_results_table <- districts_results_table |> 
  select(DISTRICT, ch_hull_Diff = chDiff)

ch_frags <- ch_frags |> 
  left_join(districts_results_table, by=c("chDISTRICT" = "DISTRICT")) |> 
  mutate(minorRatio = awMinor / awTot,
         gerry_type = case_when(
           DISTRICT == chDISTRICT ~ "packing",
           DISTRICT != chDISTRICT ~ "diluting",),
         chRatio = case_when(
           gerry_type == "packing" ~ chDiff / 100,
           gerry_type == "diluting" ~ ch_hull_Diff / 100,
           .default = 0),
         gerry = case_when(
           gerry_type == "packing" & chRatio > 0 & minorRatio - chRatio > 0 ~ (minorRatio - chRatio) * awTot,
           gerry_type == "diluting" & chRatio < 0 & minorRatio + chRatio > 0 ~ (chRatio + minorRatio) * awTot,
           .default = 0
         ),
         gerryPortionMinor = gerry / awMinor,
         gerryProportionTotal = gerry / awTot,
         gerryDots = as.integer(round(gerry/20, digits=0)),
         centroid_geom = st_centroid(geom)
         ) |> 
  dplyr::filter(gerry > 0)
```

Map local wasted votes due to gerrymandering according to the convex hull method.

```{r wasted-viz}
tmap_mode(mode = "plot")

map_local_gerry <- ch_frags |> dplyr::filter(gerry > 0) |> 
 tm_shape(bbox = st_bbox(districts)) + 
  tm_fill(
    fill = "gerryPortionMinor"
  ) 

map_local_gerry + map_districts
```

Save the resulting local gerrymander areas.

```{r warning=FALSE}
sf_use_s2(FALSE)  #ensures that GEOS engine is used rather than S2, so that geometries are planimetrically correct
ch_frags |> st_make_valid() |> 
  st_write(here("data", "derived", "public", result_set, "ch_gerry_frags.gpkg"),
           append=FALSE)
sf_use_s2(TRUE)
```

### Centroids and Dot Density

```{r}
districts_results |> st_write(here("data", "derived", "public", result_set, "district_results_LA.gpkg"),
                              append = FALSE)
districts_results |> st_drop_geometry() |> 
  write_csv(here("data", "derived", "public", result_set, "district_results_LA.csv" ))
```

# Population-weighted mean center of districts

This should be done in a planar coordinate system, but is currently not.

```{r}
pop_points <- race_fragments |> st_centroid()

district_pop_centers <- pop_points |> 
  mutate(
    wx = st_coordinates(geom)[, "X"] * Total,
    wy = st_coordinates(geom)[, "Y"] * Total
  ) |> 
  st_drop_geometry() |> 
  group_by(DISTRICT) |> 
  summarize(sumwx = sum(wx), sumwy = sum(wy), sumw = sum(Total)) |> 
  mutate(wcx = sumwx/sumw,
         wcy = sumwy/sumw) 
```
Visualize district mean centers.

```{r}
map_pop_centers <- district_pop_centers |> 
  st_as_sf(coords = c("wcx", "wcy"), crs=4269) |> 
  tm_shape() +
    tm_dots(fill = "green", size = 1.2) +
    tm_text("DISTRICT", size = 1,
            just=c("center", "top"))

map_districts + map_pop_centers
```

Calculate a distance matrix

```{r}
dcenters <- district_pop_centers |> st_drop_geometry() |> 
  select(district_2 = DISTRICT, wcx, wcy)
race_fragment_distances <- race_fragments |> 
  cross_join(dcenters) 

race_fragment_distances$dcenter <- race_fragment_distances |> 
  st_drop_geometry() |> 
  st_as_sf(coords = c("wcx", "wcy"), crs=4269) |> 
  st_geometry() 

race_fragment_distances <- race_fragment_distances |> 
  mutate(distance = st_distance(st_centroid(geom), dcenter, 
                                by_element = TRUE))
```

Visualize distance from each district's weighted population center to the locations in that district.
Very far distances suggest 

```{r}
dist <- race_fragment_distances |> st_drop_geometry() |> 
  dplyr::filter(DISTRICT == district_2) |> 
  select(frag_id, distance)

race_fragments |> left_join(dist, by="frag_id") |> 
  tm_shape() +
  tm_fill(fill = "distance",
          fill.scale = tm_scale_intervals(n = 10, style = "equal", values = "viridis")) +
  map_districts +
  map_pop_centers
```

Construct compact districts

- `idealpop` is the total voting age population divided by the number of districts. Notably, this is not the same as the total population, which is the criteria used in redistricting. This could be revisited, and would require areal interpolation of total population.
- `compact_pop_center_fragments` are the population fragments belonging to each compact population-weighted center district
- `compact_pop_center_districts` are the aggregated districts based on the compact population-weighted center

```{r}
idealpop <- sum(blockgroups_calc$Total)/6

compact_pop_center_fragments <- race_fragment_distances |> 
  st_drop_geometry() |> 
  arrange(district_2, distance) |> 
  group_by(district_2) |> 
  mutate(cumulative_pop = cumsum(Total)) |> 
  dplyr::filter(cumulative_pop <= idealpop) |> 
  ungroup() |> 
  select(frag_id, cpcDistrict = district_2) |> 
  left_join(race_fragments, by = "frag_id") |> 
  st_set_geometry("geom")

compact_pop_center_districts <- compact_pop_center_fragments |> 
  group_by(cpcDistrict) |> 
  summarize(cpcMinority = sum(minority),
            cpcTotal = sum(Total),
            geom = st_union(geom)) 
```

Visualize the compact population-weighted center districts

```{r warning=FALSE}
map_districts +
compact_pop_center_districts |> st_buffer(0.0000001) |> st_buffer(-0.0000001) |> 
  mutate(District_Number = as.character(cpcDistrict)) |> 
  tm_shape() +
  tm_polygons(fill = "District_Number",
              fill_alpha = 0.3)
```

### District Cdiff scores based on compact population centers

Join convex hull estimates to Districts with block group estimates.

- `cpcDiff` is the District composition minus the compact population center composition. A positive number indicates over-representation (packing?) in the district compared to the convex hull of the district, while a negative number represents an under-representation (dilution? cracking?) in the district compared to the convex hull of the district.
- `cpcAbsDiff` is the absolute value of cpcDiff to measure total change in composition of the district compared to the convex hull of the district.

```{r}
d_cpc_est <- compact_pop_center_districts |> 
  st_drop_geometry() 

districts_results <- districts_results |> 
  left_join(d_cpc_est, by = c("DISTRICT" = "cpcDistrict")) |> 
  mutate(cpcPct = round(cpcMinority / cpcTotal * 100, 1),
         cpcDiff = pctMinoritybg - cpcPct,
         cpcAbsDiff = abs(cpcDiff)) 
```

## Calculate local gerrymandering based on cpc Districts

### Local Gerrymandering Scores

- Join District-level `cpcDiff` data to the `compact_pop_center_fragments` fragments from overlaying convex hulls with census geometries.
- `minorRatio` is the local minority ratio of the fragment
- `gerry_type` is the type of gerrymandering for the given fragment: `packing` when the fragment is inside of its district (`DISTRICT == chDISTRICT`) and `diluting` when he fragment is outside of its district (`DISTRICT != chDISTRICT`)
- `chRatio` is the minority ratio of the relevant district or compact district shape, which is the chDiff of the district for packing and the chDiff of the compact district (`cpc_compact_Diff`) for diluting.
- `gerry` represents total wasted minority votes due to gerrymandering, estimated in the packing form as the proportion of voting-age minorities in the fragment in excess of the proportion of voting-age minorities in the convex hull, and in the dilution form as the proportion of voting-age minorities in the fragment in excess of the proportion of voting-age minorities in the district.

**CAUTION:** I have relied too much on comparing the district #'s and the assumption with convex hull that the district is completely inside of the compact shape. Therefore, I am missing some information in the local gerrymandering scores, e.g. the part of a packed district that is outside of the compact shape, and the part of a diluted district that is inside its compact shape.

```{r wasted-votes}
districts_results_table <- districts_results |> st_drop_geometry() |> 
  select(DISTRICT, cpcDiff)

# frag_cpc_dist <- compact_pop_center_fragments |> 
#   st_drop_geometry() |>
#   select(frag_id, cpcDistrict)

frag_identity <- compact_pop_center_fragments |> 
  st_drop_geometry() |> 
  filter(DISTRICT == cpcDistrict) |> 
  select(frag_id)

frag_orphans <- race_fragments |> 
  anti_join(frag_identity, by="frag_id")

cpc_frags <- compact_pop_center_fragments |> 
  bind_rows(frag_orphans) |> 
  left_join(districts_results_table, by="DISTRICT") |> 
  left_join(select(districts_results_table, 
                   DISTRICT, 
                   cpc_compact_Diff = cpcDiff),
            by=c("cpcDistrict" = "DISTRICT"))

cpc_frags <- cpc_frags |> 
  mutate(minorRatio = minority / Total,
         gerry_type = case_when(
           DISTRICT == cpcDistrict ~ "packing",
           is.na(cpcDistrict) & cpcDiff > 0 ~ "packing",
           DISTRICT != cpcDistrict ~ "diluting",),
         cpcRatio = case_when(
           gerry_type == "packing" ~ cpcDiff / 100,
           gerry_type == "diluting" ~ cpc_compact_Diff / 100,
           .default = 0),
         gerry = case_when(
           gerry_type == "packing" & cpcRatio > 0 & minorRatio - cpcRatio > 0 ~ (minorRatio - cpcRatio) * Total,
           gerry_type == "diluting" & cpcRatio < 0 & minorRatio + cpcRatio > 0 ~ (cpcRatio + minorRatio) * Total,
           .default = 0
         ),
         gerryPortionMinor = gerry / minority,
         gerryProportionTotal = gerry / Total,
         gerryDots = as.integer(round(gerry/20, digits=0)),
         centroid_geom = st_centroid(geom)
         ) 

# |>  dplyr::filter(gerry > 0)
```

Map local wasted votes due to gerrymandering according to the compact weighted population center method.

```{r wasted-viz}
tmap_mode(mode = "plot")

map_local_gerry <- cpc_frags |> 
  dplyr::filter(gerry > 0) |> 
 tm_shape(bbox = st_bbox(districts)) + 
  tm_fill(
    fill = "gerry_type",
    fill_alpha = 0.3
  ) 

map_local_gerry + map_districts
```

Something to consider: since the compact weighted population centers leave some portions of the state uncovered,
do minority votes in those areas constitute some form of wasted vote in the dilution case?
For Louisiana specifically, sections of the border between districts 3 and 4 and districts 4 and 5 are contorted to allocate minority votes in areas that are not covered by any compact weighted population center district.

Further, with this population-based approach, it would also be informative to simply find the portion of population of a district that is included in the compact mean population center shape.

# Results

### Correlation matrix and small plots for gerrymandering indicators

```{r}
districts_results_cor <- districts_results |> 
  st_drop_geometry() |> 
  select(pctMinoritybg,
      #   chPct,
         chAbsDiff,
      #   mbcPct,
         mbcAbsDiff,
         compact_shp,
         compact_hull,
         compact_circ)

districts_results_cor |> cor() |> corrplot(type="upper", diag=FALSE, method="number") 

#districts21_results_cor |> cor() |> kable()
```

```{r}
districts_results_cor <- districts_results |> 
  st_drop_geometry() |> 
  select(pctMP,
      #   chPct,
         chAbsDiffV,
      #   mbcPct,
         mbcAbsDiffV,
         compact_shp,
         compact_hull,
         compact_circ)

districts_results_cor |> cor() |> corrplot(type="upper", diag=FALSE, method="number")

#districts21_results_cor |> cor() |> kable()
```


Follow the first percent Black row across.
While it is most correlated with the minimum bounding circle approach to compactness,
there are also strong positive correlations with the degree of change in the Percentage of the population that is a minority between the district and its convex hull or its minimum bounding circle.
In other words, if you change the shape of a district into a convex hull or minimum bounding circle,
the composition of minority-majority districts changes more than the composition of majority-majority districts.
Meanwhile, there is little correlation with how compact a shape is.


### Plot representational difference against compactness

Scatterplot with (absolute) difference in representation on x axis and compactness on y axis.
Plot the three different compactness scores simultaneously with different colors.
Symbolize the districts with different shapes.

```{r}
districts_results_plot <- districts_results |> 
  st_drop_geometry() |> 
  select(DISTRICT, pctMinoritybg, chAbsDiff, compact_shp, compact_hull, compact_circ) |> 
  pivot_longer(cols = starts_with("compact"))

districts_results_plot |> ggplot() +
  aes(x = chAbsDiff, y = value) +
  geom_smooth(method="lm", col = "grey30") +
  geom_label(aes(label = DISTRICT, fill = pctMinoritybg)) +
  scale_fill_distiller(type = "div", palette = "PRGn") +

  facet_wrap(~name)
```



There is a negative relationship between convex hull compactness and convex hull difference.
There is a negative relationship between convex hull compactness and convex hull difference.
There is a negative relationship between minimum bounding circle compactness and convex hull representational difference.  
The exceptions are districts 5 and 7. 
District 7 really is gerrymandered (packed African American), but the minimum bounding circle method does not find it so.
District 5 is not really gerrymandered, even though the minimum bounding circle does find it so.

Shape and convex hull exhibit a positive correlation.  
Shape and minimum bounding circle exhibit a positive correlation, with the exception of District 5.
Convex hull and minimum bounding circle exhibit a positive correlation, with the exception of District 5.

District 5 is a long, but otherwise compact shape.

```{r}
tm_shape(districts_results) +
  tm_polygons(fill = "pctMinoritybg") +
  tm_text("DISTRICT")
```


# Discussion

Describe how the results are to be interpreted *vis a vis* each hypothesis or research question.


# Acknowledgements

This report is based upon the template for Reproducible and Replicable Research in Human-Environment and Geographical Sciences, DOI: [10.17605/OSF.IO/W29MQ](https://doi.org/10.17605/OSF.IO/W29MQ)

# References
