---
title: "`r paste0('A Parsimonious Approach to Gerrymandering Analysis (', params$state, ')')`"
author:
  - "1. Joseph Holler (josephh@middlebury.edu; corresponding author), Middlebury College"
  - "2. Peter Kedron (peterkedron@ucsb.edu), University of California, Santa Barbara"
  - "3. Pyung Kim (pyungkim@ucsb.edu), University of California, Santa Barbara"
date: "`r Sys.Date()`"
knit: (function(input, encoding) { p <- rmarkdown::yaml_front_matter(input)$params; rmarkdown::render(input, encoding = encoding, output_file = paste0("Gerrymander_", gsub("[^A-Za-z0-9]+", "_", paste(p$state, p$group, p$district_year)), ".html")) })
output:
  bookdown::html_document2:
    code_folding: hide
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: false
    fig_caption: true
editor_options:
  markdown:
    wrap: sentence
params:
  # Set `state`, `group`, and `district_year`, then Knit. Two steps are not automatic.
  # A free Census API key must be installed once,
  # and the precinct election file must be downloaded by hand.
  # See "How to run this analysis" below.
  state: "Tennessee"    # the state to analyze; type any U.S. state name here
  group: "Black"        # group to analyze: a P3 race (Black, Asian, ...) or "Hispanic"
  district_year: 2024   # congressional map year from TIGER; match your research question
---

```{=html}
<style>
/* more space before a higher-level heading, less before a lower one, and tight
   spacing inside a section so one numbered part reads as a single block */
h1 { margin-top: 3.4em; margin-bottom: 0.4em; }
h2 { margin-top: 2.2em; margin-bottom: 0.3em; }
h3 { margin-top: 1.3em; margin-bottom: 0.3em; }
h4 { margin-top: 0.8em; margin-bottom: 0.2em; }
p  { margin-top: 0.2em; margin-bottom: 0.5em; }
.figure, img { margin-top: 0.4em; margin-bottom: 0.4em; }

/* figure captions ("Figure N: ...") large and dark so they read as titles */
.figure .caption, div.figure p.caption, p.caption, .caption {
  font-size: 1.5rem !important;
  font-weight: 700 !important;
  color: #16232a !important;
  margin-top: 0.6em;
  text-align: left;
  line-height: 1.35;
}
/* keep the table body compact without shrinking the caption (font_size in
   kable_styling would force the caption to a fixed size and block styling) */
.table, table.table { font-size: 14px; }
/* table captions sit above the table, prominent rather than faint grey.
   rem keeps them root-relative so the 14px table body does not shrink them */
table > caption, .table > caption, table caption, caption {
  caption-side: top !important;
  font-size: 1.5rem !important;
  font-weight: 700 !important;
  color: #16232a !important;
  text-align: left !important;
  padding-bottom: 0.4em;
  line-height: 1.35;
}
/* the "Table N:" label bold (kableExtra wraps it in a span carrying the tab id) */
caption span[id^="tab:"] { font-weight: 700; }
</style>
```

# Abstract {.unnumbered}

This is a study of gerrymandering in the United States. We introduce a parsimonious
composite approach that compares each district's racial and partisan composition to
that of compact reference shapes. We implement the entire workflow so that it can be
applied to any U.S. state from standard, publicly available data.

## Study Metadata {.unnumbered}

- `Key words`: gerrymandering, compactness, convex hull, racial and partisan representation
- `Subject`: Social and Behavioral Sciences: Geography: Geographic Information Sciences
- `Date created`: 2025-02-17
- `Date modified`: `r Sys.Date()`
- `Spatial Coverage`: United States; one state per run, selected via `params$state` (this run: `r params$state`)
- `Spatial Resolution`: Census block groups (demographics) and voting precincts (elections)
- `Spatial Reference System`: EPSG:4269 NAD 1983 Geographic Coordinate System
- `Temporal Coverage`: 2020 decennial census population; 2024 presidential election results
- `Temporal Resolution`: Decennial census

# 1 Study design

## 1.1 Overview

This is an original, exploratory study based on the literature on gerrymandering
metrics. We evaluate the usefulness of a new gerrymandering metric that measures how
much a congressional district's representativeness differs from that of compact
reference shapes derived from it. These reference shapes are its convex hull, its
minimum bounding circle, and a compact district grown from its population-weighted
mean center.

## 1.2 Bias and threats to validity

This study is explicitly an investigation into the modifiable areal unit problem.
Shape-based measures are sensitive to edge effects and scale, because complex borders
formed by natural features such as coastlines and rivers vary greatly in perimeter
depending on the scale of analysis. We hope this study establishes a method that is
more robust to the threats to validity caused by scale and edge effects in studies of
gerrymandering and district shape.

# 2 How to run this analysis

1. Set the analysis in the header. `params$state` is the state, `params$group` is the
   racial or ethnic group to analyze (a race such as "Black" or "Asian", or "Hispanic"),
   and `params$district_year` is the congressional map year.
2. Choose `params$district_year` for your research purpose. To evaluate a particular
   election, use the year of the map that election was run under, since some states redrew
   their map mid-decade. You can also override TIGER entirely by placing a specific plan as
   `districts.gpkg` in the state folder `data/raw/public/<state>/`.
3. Get a free Census API key at https://api.census.gov/data/key_signup.html and
   install it once by running `census_api_key("YOUR KEY", install = TRUE)` in the
   RStudio console, then restart R. This lets the document download census block
   groups and congressional districts. You can skip it if you supply those as local
   files.
4. To include the voting layer, download that state's "2024 General Election Precinct
   Boundaries and Election Results" from the Redistricting Data Hub and unzip it into
   `data/raw/public/<state>/`. The code finds the `_all_` precinct shapefile inside that folder
   automatically, so you can leave the downloaded folder as is. The congressional
   districts and census block groups download automatically, so you do not need those.
5. Knit the document. The code reads `params$state` and loads that state's data.

# 3 Materials and procedure

## 3.1 Computational environment

We use `tidyverse`, `sf`, and `lwgeom` for data wrangling and spatial operations. We
use `tidycensus` and `tigris` to obtain census demographics and district geographies. We
use `ggplot2` (part of `tidyverse`) for mapping and `knitr` with `kableExtra` for tables.
The setup chunk installs any missing package so the analysis runs on a clean machine.

```{r environment-setup, include = FALSE}
options(scipen = 999, digits = 3)

# packages used in this analysis
packages <- c("tidyverse", "here", "sf", "lwgeom",
              "tidycensus", "tigris", "knitr", "kableExtra", "ggtext", "bookdown")

# install any that are missing, then load them
for (p in packages) {
  if (!requireNamespace(p, quietly = TRUE))
    install.packages(p, repos = "https://cloud.r-project.org")
  suppressPackageStartupMessages(library(p, character.only = TRUE))
}

# use the GEOS planar engine rather than the s2 spherical engine. Precinct files
# from different states vary in quality, and s2 halts on tiny geometry defects
# such as duplicate vertices. GEOS repairs and measures them without stopping.
sf::sf_use_s2(FALSE)
```

```{r rec-environment, warning = FALSE, message = FALSE}
# default knitr options
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.width = 8,
                      warning = FALSE, message = FALSE)

# one consistent table style used throughout, with a colored bold header row
style_tbl <- function(kb) {
  kb |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width = TRUE, position = "center") |>
    kableExtra::row_spec(0, bold = TRUE, color = "white", background = "#3B6B7A")
}

# color for Cdiff cells, matching the maps and bars: packing (positive) red,
# cracking (negative) blue, near-zero grey
diff_color <- function(x) ifelse(x > 0, "#C0405E", ifelse(x < 0, "#3E7EA0", "#555555"))

# one color per reference shape, used to tint the facet strip labels (figures)
# and the grouped column headers (tables): hull green, circle pink, center orange
ref_col <- c("Convex Hull" = "#3E9B63", "Min. Bounding Circle" = "#7D3C98",
             "Compact Pop. Center" = "#D98400")
ref_labels <- setNames(
  paste0("<span style='color:", ref_col, "'>", names(ref_col), "</span>"),
  names(ref_col))

# grouped column header for the results tables (plain, no color)
ref_header <- c(" " = 2, "Convex hull" = 2, "Min. bounding circle" = 2,
                "Compact pop. center" = 2)

# optionally record the R session and package citations; never halt the knit if
# the folders are missing or not writable
try(silent = TRUE, {
  dir.create(here("procedure", "environment"), recursive = TRUE, showWarnings = FALSE)
  writeLines(capture.output(sessionInfo()),
             here("procedure", "environment", paste0("r-environment-", Sys.Date(), ".txt")))
  knitr::write_bib(c(packages, "base"), file = here("software.bib"))
})
```

## 3.2 Data and variables

### 3.2.1 Data sources

The workflow reads each dataset from a local file when present, and automatically
downloads the ones that have a programmatic source for `params$state`. Only the
precinct data must be downloaded by hand.

- **Congressional districts.** The boundaries of the districting plan under study.
  Downloaded automatically from the U.S. Census TIGER files with the `tigris`
  package, or read from a local plan file obtained from the state legislature or the
  Redistricting Data Hub.
- **Voting precincts with election results.** Precinct boundaries joined to
  general-election returns, used for the partisan analysis. There is no public API,
  so this file is downloaded by hand from the Redistricting Data Hub at
  https://redistrictingdatahub.org (free account). On the state page, choose the
  dataset named "General Election [State] VTD (Precinct) Level Results and Boundaries"
  for the year of interest (for example 2024) and download the shapefile. Vote
  columns follow the `G<yr>PRE<party><cand>` naming convention, for example
  `G24PRERTRU` for 2024 presidential votes for the Republican candidate.
- **Census block groups.** Population 18 years and older by race, from the 2020
  Census PL 94-171 redistricting file, table `P3`, downloaded automatically with the
  `tidycensus` package. Block group geometry is included in the same download.

The `P3` variable metadata that labels which columns count as the minority group is a
supporting lookup. It is downloaded automatically with `tidycensus::load_variables`
and cached as `census2020pl_vars.csv`.

### 3.2.2 Loading the data

We set the file paths from `params$state` and load each dataset, downloading the ones
with a programmatic source when a local copy is not present.

```{r load-data, results = "hide"}
# per-state data folder and file paths (edit names to match your files)
state_slug      <- gsub("[^A-Za-z0-9]+", "_", params$state)
data_dir        <- here("data", "raw", "private", state_slug)
output_dir      <- here("data", "derived", "private", state_slug)
districts_file  <- here(data_dir, "districts.gpkg")   # a plan file, or fetched from TIGER
blockgroup_file <- here(data_dir, "block_groups.gpkg")
census_metadata_file <- here("data", "metadata", "census2020pl_vars.csv")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

# precincts: use the Redistricting Data Hub "_all_" precinct shapefile found anywhere
# under the state folder, or a file named precincts.shp; fall back if none is present
precincts_file  <- {
  hits <- list.files(data_dir, pattern = "_all_prec\\.shp$", recursive = TRUE, full.names = TRUE)
  if (!length(hits)) hits <- list.files(data_dir, pattern = "precincts\\.shp$", recursive = TRUE, full.names = TRUE)
  if (length(hits)) hits[1] else file.path(data_dir, "precincts.shp")
}

# make sure the folders used to cache downloaded data exist
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(census_metadata_file), recursive = TRUE, showWarnings = FALSE)

# districts: a local plan file if present, otherwise the enacted districts from TIGER
if (file.exists(districts_file)) {
  districts <- st_read(districts_file)
} else {
  districts <- tigris::congressional_districts(state = params$state, cb = TRUE,
                                               year = params$district_year)
}

# precincts with election results: local file only (manual download), optional
have_votes <- file.exists(precincts_file)
if (have_votes) precincts <- st_read(precincts_file)

# census variable metadata: cached csv, or query and cache
if (file.exists(census_metadata_file)) {
  census2020pl_vars <- read.csv(census_metadata_file)
} else {
  census2020pl_vars <- load_variables(2020, "pl")
  write.csv(census2020pl_vars, census_metadata_file)
}

# census block groups: P3 (race for 18+), P4 (Hispanic ethnicity for 18+), and P1_001N
# (total population all ages); cached, or queried and cached
if (file.exists(blockgroup_file)) {
  blockgroups <- st_read(blockgroup_file)
} else {
  blockgroups <- get_decennial(geography = "block", sumfile = "pl", table = "P3",
                               year = 2020, state = params$state, output = "wide",
                               geometry = TRUE, keep_geo_vars = TRUE)
  # Hispanic or Latino ethnicity (table P4), for the group = "Hispanic" option
  hisp <- get_decennial(geography = "block", sumfile = "pl", table = "P4",
                        year = 2020, state = params$state, output = "wide")
  # total population of all ages, used later to size the compact districts
  totalpop <- get_decennial(geography = "block", sumfile = "pl",
                            variables = "P1_001N", year = 2020, state = params$state,
                            output = "wide")
  blockgroups <- blockgroups |>
    left_join(dplyr::select(hisp, GEOID, dplyr::starts_with("P4_")), by = "GEOID") |>
    left_join(dplyr::select(totalpop, GEOID, P1_001N), by = "GEOID")
  st_write(blockgroups, blockgroup_file)
}

# if a cached file predates Hispanic support, fetch table P4 on demand (also used by the
# full race and ethnicity composition table)
if (!"P4_002N" %in% names(blockgroups)) {
  hisp <- get_decennial(geography = "block group", sumfile = "pl", table = "P4",
                        year = 2020, state = params$state, output = "wide")
  blockgroups <- left_join(blockgroups, dplyr::select(hisp, GEOID, dplyr::starts_with("P4_")),
                           by = "GEOID")
}
```

## 3.3 Data processing

### 3.3.1 Spatial transformations

The districts, precincts, and block groups arrive in different coordinate reference
systems. To perform spatial operations we transform them all to a common CRS, the
NAD 1983 geographic coordinate system (EPSG:4269). We also standardize the district
identifier column, whose name varies by data source, to `DISTRICT`.

```{r transform}
districts <- st_make_valid(st_transform(districts, 4269))
names(districts)[names(districts) == attr(districts, "sf_column")] <- "geom"
st_geometry(districts) <- "geom"

# standardize the district id column to "DISTRICT" (names vary by source)
dcol <- c("DISTRICT", "DISTRICT_I", grep("^CD[0-9]+FP$", names(districts), value = TRUE))
dcol <- dcol[dcol %in% names(districts)][1]
districts$DISTRICT <- as.integer(as.character(districts[[dcol]]))

blockgroups <- st_make_valid(st_transform(blockgroups, 4269))
names(blockgroups)[names(blockgroups) == attr(blockgroups, "sf_column")] <- "geom"
st_geometry(blockgroups) <- "geom"

if (have_votes) {
  precincts <- st_make_valid(precincts) |> st_transform(4269)
  names(precincts)[names(precincts) == attr(precincts, "sf_column")] <- "geom"
  st_geometry(precincts) <- "geom"
}

# state aspect ratio (height / width), used to size maps so they do not leave whitespace
.bb <- st_bbox(districts)
map_w   <- 8   # map width in inches
map_asp <- as.numeric((.bb["ymax"] - .bb["ymin"]) /
                      ((.bb["xmax"] - .bb["xmin"]) * cospi(mean(c(.bb["ymin"], .bb["ymax"])) / 180)))
```


```{r map-districts, echo = FALSE, fig.width = map_w, fig.height = map_w * map_asp, fig.cap = paste0("Congressional districts of ", params$state, ", each labeled by its congressional district number.")}
ggplot(districts) +
  geom_sf(fill = "#C1E3CD", color = "#006d2c", alpha = 0.5) +
  geom_sf_text(aes(label = DISTRICT), color = "black", size = 5, fontface = "bold") +
  theme_void()
```

### 3.3.2 Compute variables of interest

We compute the percentage of the population 18 years and older that belongs to the group
named by `params$group`. For a race such as Black or Asian we select the `P3` variables
whose label names that group, counting anyone who identifies with that category alone or
in combination with any other race. For Hispanic or Latino, which the census treats as an
ethnicity rather than a race, we use the `P4` table instead.

```{r minority-vars}
group_label <- params$group   # a P3 race (e.g. Black, Asian) or "Hispanic" from table P4

if (tolower(group_label) == "hispanic") {
  # Hispanic or Latino, population 18 and over (table P4, variable P4_002N)
  minority_vars <- census2020pl_vars |>
    dplyr::filter(name == "P4_002N") |>
    dplyr::select(name, label)
} else {
  # a race category, population 18 and over (table P3, alone or in combination)
  minority_vars <- census2020pl_vars |>
    dplyr::filter(str_detect(name, "^P3"), str_detect(label, group_label)) |>
    dplyr::select(name, label)
}
```

```{r table1, echo = FALSE}
minority_vars |>
  rename(Variable = name, Label = label) |>
  kbl(caption = "Selected census variables for the group of interest.") |>
  style_tbl() |>
  scroll_box(height = "300px")
```

For each block group we sum the selected columns into `minority`, copy the 18-and-over
total (`P3_001N`) into `Total`, and compute `pctMinority`. `CheckPct` is the share of
the population that is either White alone or the minority group, which should be near
100 percent unless other minority groups are large. We also compute the area of each
block group for later area-weighted reaggregation.

```{r bg-vars}
# fall back to voting-age population if total population is absent (older cached file)
if (!"P1_001N" %in% names(blockgroups)) blockgroups$P1_001N <- blockgroups$P3_001N

blockgroups_calc <- blockgroups |>
  dplyr::filter(!st_is_empty(geom)) |>
  rowwise() |>
  mutate(minority = sum(c_across(all_of(minority_vars$name)))) |>
  ungroup() |>
  mutate(sarea = st_area(geom),
         Total = P3_001N,     # voting-age population (18+), used for composition
         TotalPop = P1_001N,  # all ages, used to size the compact districts
         pctMinority = minority / Total * 100,
         CheckPct = (minority + P3_003N) / Total * 100) |>
  dplyr::select(GEOID, sarea, minority, Total, TotalPop, pctMinority, CheckPct)
```

For the precincts we compute `totV`, the two-party presidential total, and `mpV`, the
minority party vote. The majority and minority party designations are state specific,
so we choose the minority party automatically as the statewide underdog, the party
with fewer total votes.

```{r precinct-vars}
if (have_votes) {
  # 2024 two-party presidential columns, detected by name
  # (Republican G24PRER*, Democratic G24PRED*)
  rep_cols <- grep("^G24PRER", names(precincts), value = TRUE)
  dem_cols <- grep("^G24PRED", names(precincts), value = TRUE)

  Rtot <- sum(rowSums(st_drop_geometry(precincts)[rep_cols]), na.rm = TRUE)
  Dtot <- sum(rowSums(st_drop_geometry(precincts)[dem_cols]), na.rm = TRUE)
  minority_party <- if (Dtot <= Rtot) "Democratic" else "Republican"
  mp_cols <- if (minority_party == "Democratic") dem_cols else rep_cols

  precincts <- precincts |>
    mutate(totV = rowSums(across(all_of(c(rep_cols, dem_cols)))),
           mpV  = rowSums(across(all_of(mp_cols))),
           sarea = st_area(geom))
}
```


```{r disaggregate-votes}
if (have_votes) {
  # Allocate each precinct's votes to intersecting block groups by voting-age population.
  vote_fragments <- st_intersection(
    precincts |>
      mutate(precinct_id = dplyr::row_number()) |>
      dplyr::select(precinct_id, totV, mpV),
    blockgroups_calc |>
      dplyr::select(GEOID, Total, sarea)
  ) |>
    mutate(bg_weight = Total * as.numeric(st_area(geom) / sarea),
           prec_totV = totV,
           prec_mpV = mpV) |>
    group_by(precinct_id) |>
    mutate(weight = bg_weight / sum(bg_weight),
           totV = prec_totV * weight,
           mpV = prec_mpV * weight) |>
    ungroup() |>
    group_by(GEOID) |>
    summarize(totV = sum(totV), mpV = sum(mpV), .groups = "drop")

  blockgroups_calc <- blockgroups_calc |>
    left_join(st_drop_geometry(vote_fragments), by = "GEOID") |>
    mutate(totV = coalesce(totV, 0),
           mpV = coalesce(mpV, 0))
}
```


### 3.3.3 Descriptive statistics (racial and ethnic composition by district)

Table \@ref(tab:desc-race) summarizes the racial and ethnic composition of the voting-age
population before the analysis. The first row is the statewide baseline, and each following row is a district.
District shares come from the same area-weighted reaggregation of the block groups used in
the rest of the analysis. The race categories are counted alone and add up to about 100
percent. To read the tables, compare each district row to the Statewide row on top. A district
whose share is far above the statewide share holds a lot of that group, and one far below holds
little. Big swings between districts are the raw material the rest of the analysis examines,
since gerrymandering works by concentrating a group in a few districts (packing) or spreading it
thin across many (cracking).

# Disabled section

Disabled the following code blocks, which need to be:

- consistent with disaggregating voting data into census blocks
- responsive to a decision to be inclusive or exclusive with regards to intersectional racial/ethnic identities
- comprehensive: just summarize all data from blocks simulaneously in one descriptive table


```
race_vars <- c(White = "P3_003N", Black = "P3_004N",
               `Am. Indian / AK Native` = "P3_005N", Asian = "P3_006N",
               `Nat. Hawaiian / Pac. Isl.` = "P3_007N", `Some other race` = "P3_008N",
               `Two or more races` = "P3_009N")
eth_vars <- c(`Hispanic or Latino` = "P4_002N", `Not Hispanic or Latino` = "P4_003N")
vote_vars <- c(`Total Votes` = "totV", `Minority Party Votes` = "mpV")
all_vars <- c(race_vars, eth_vars, vote_vars)

# per-district composition by area-weighted reaggregation of the block groups
bg_race <- blockgroups |>
  dplyr::filter(!st_is_empty(geom)) |>
  mutate(sarea = st_area(geom)) |>
  dplyr::select(sarea, Total = P3_001N, dplyr::all_of(unname(all_vars)))

dist_comp <- st_intersection(bg_race, districts) |>
  st_make_valid() |> st_collection_extract("POLYGON") |>
  dplyr::filter(!st_is_empty(geom)) |>
  mutate(w = as.numeric(st_area(geom) / sarea)) |>
  st_drop_geometry() |>
  group_by(DISTRICT) |>
  summarize(dplyr::across(c(Total, dplyr::all_of(unname(all_vars))), ~ sum(.x * w)),
            .groups = "drop") |>
  arrange(DISTRICT)

# statewide totals
sw <- blockgroups |> st_drop_geometry() |>
  dplyr::summarize(Total = sum(P3_001N),
                   dplyr::across(dplyr::all_of(unname(all_vars)), ~ sum(.x)))

# percentages, statewide row on top
pct <- bind_rows(sw, dplyr::select(dist_comp, -DISTRICT)) |>
  mutate(dplyr::across(dplyr::all_of(unname(all_vars)), ~ round(.x / Total * 100, 1)))
district_labels <- c("Statewide", as.character(dist_comp$DISTRICT))

race_tab <- cbind(District = district_labels, dplyr::select(pct, dplyr::all_of(unname(race_vars))))
names(race_tab) <- c("District", names(race_vars))
race_tab |>
  kbl(caption = "Race composition of the voting-age population by district (%).") |>
  style_tbl() |>
  row_spec(1, bold = TRUE, background = "#EAF0F1") |>
  scroll_box(width = "100%")
```

Hispanic or Latino is an ethnicity that overlaps with race, so it is reported in a separate
table (Table \@ref(tab:desc-hispanic)) and is not part of the sum above.

```
hisp_tab <- cbind(District = district_labels, dplyr::select(pct, dplyr::all_of(unname(eth_vars))))
names(hisp_tab) <- c("District", names(eth_vars))
hisp_tab |>
  kbl(caption = "Hispanic or Latino ethnicity of the voting-age population by district (%).") |>
  style_tbl() |>
  row_spec(1, bold = TRUE, background = "#EAF0F1")
```

### 3.3.4 Dot density map (where the group lives)

A dot-density map (Figure \@ref(fig:map-dots-group)) shows where people live, which the tables above cannot show. Each dot is a
fixed number of people, the group set by `params$group` in one color and everyone else in grey.
This is a descriptive picture, not a result, but it matters because packing and cracking depend on
where people already live. To read it, look for tight clusters of the group's color, which mark
where it concentrates, often a city or a historic community, and compare those clusters to the
district lines. When a district line wraps tightly around a cluster, that is a visual sign of
packing, and when a line cuts a cluster into several districts, that is a visual sign of cracking.


# Disabled section

Dot density mapping becomes infeasible with large data and small enumeration areas.

```
set.seed(1)
dot_people <- 1000   # people per dot (voting age); raise this to speed up rendering
bg_dots <- blockgroups |> dplyr::filter(!st_is_empty(geom))

# helper: build one point layer per category column, sampled in proportion to its count
sample_dots <- function(bg, cols) {
  parts <- lapply(names(cols), function(nm) {
    n   <- round(pmax(st_drop_geometry(bg)[[cols[[nm]]]], 0) / dot_people)
    sfc <- st_sample(bg, size = n)
    if (length(sfc) == 0) NULL else st_sf(kind = nm, geometry = sfc)
  })
  out <- do.call(rbind, parts[!vapply(parts, is.null, logical(1))])
  out$kind <- factor(out$kind, levels = names(cols))
  out
}
```

```
{map-dots-group, eval = FALSE,  echo = FALSE, fig.width = map_w, fig.height = map_w * map_asp + 0.6, fig.cap = paste0("Dot density of ", params$group, " and everyone else, ", params$state, ".")}
if (tolower(params$group) == "hispanic") {
  bg_dots$grp <- bg_dots$P4_002N
} else {
  bg_dots$grp <- rowSums(st_drop_geometry(bg_dots)[minority_vars$name], na.rm = TRUE)
}
bg_dots$oth <- pmax(bg_dots$P3_001N - bg_dots$grp, 0)

ggplot() +
  geom_sf(data = sample_dots(bg_dots, setNames(c("grp", "oth"), c(params$group, "Other"))),
          aes(color = kind), size = 0.15, alpha = 0.5) +
  geom_sf(data = districts, fill = NA, color = "grey40", linewidth = 0.3) +
  scale_color_manual(values = setNames(c("#2166ac", "grey75"), c(params$group, "Other")),
                     name = NULL) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(subtitle = paste0("1 dot = ", dot_people, " people (voting age)")) +
  theme_void(base_size = 14) +
  theme(legend.position = "bottom", plot.margin = margin(0, 0, 0, 0))
```

# 4 Analysis

## 4.1 District composition by area-weighted reaggregation

We estimate the minority share of each district by area-weighted reaggregation of the
block groups. Intersecting the block groups with the districts splits each block group
into fragments, one per district it touches. We allocate each block group's population
to its fragments in proportion to area, then sum the fragments by district. We keep the
fragments (`race_fragments`) because later steps reuse them. The intersection is made
robust by keeping only polygons, repairing invalid geometries, dropping empty results,
and casting to single-part polygons.

```{r district-race, warning = FALSE}
race_fragments <- st_intersection(blockgroups_calc, districts) |>
  st_make_valid() |>
  st_collection_extract("POLYGON") |>   # drop line/point slivers from planar overlaps
  dplyr::filter(!st_is_empty(geom)) |>
  st_cast("MULTIPOLYGON") |> st_cast("POLYGON") |>
  mutate(awTot    = Total * as.numeric(st_area(geom) / sarea),
         awMinor  = minority * as.numeric(st_area(geom) / sarea),
         awTotPop = TotalPop * as.numeric(st_area(geom) / sarea),
         awtotV = totV * as.numeric(st_area(geom) / sarea),
         awmpV = mpV * as.numeric(st_area(geom) / sarea),
         frag_id  = row_number()) |>
  dplyr::select(DISTRICT, frag_id, sarea, 
                Total = awTot, 
                minority = awMinor,
                TotalPop = awTotPop,
                totV = awtotV,
                mpV = awmpV)

district_estimates <- race_fragments |>
  st_drop_geometry() |>
  group_by(DISTRICT) |>
  summarize(bgTotal = sum(Total), 
            bgMinority = sum(minority),
            totV = sum(totV),
            mpV = sum(mpV),
            .groups = "drop")

districts_join <- districts |>
  left_join(district_estimates, by = "DISTRICT") |>
  mutate(pctMinoritybg = round(bgMinority / bgTotal * 100, 1),
         pctMP = round(mpV / totV * 100, 1))
```

## 4.2 Convex hull composition

The convex hull of a district is the smallest convex shape that contains it, like a
rubber band stretched around the district. We build one hull per district and estimate
the minority share inside it by the same area-weighted reaggregation, this time
intersecting the block groups with the hulls. When precinct data is present we do the
same for votes.

```{r convex-hull, warning = FALSE}
d_ch <- districts |>
  dplyr::select(DISTRICT) |>
  st_convex_hull() |>
  mutate(chArea = st_area(geom))

d_ch_est <- st_intersection(blockgroups_calc, d_ch) |>
  st_make_valid() |>
  st_collection_extract("POLYGON") |>
  dplyr::filter(!st_is_empty(geom)) |>
  mutate(awTot   = Total * as.numeric(st_area(geom) / sarea),
         awMinor = minority * as.numeric(st_area(geom) / sarea)) |>
  st_drop_geometry() |>
  group_by(DISTRICT, chArea) |>
  summarize(chTotal = sum(awTot), chMinority = sum(awMinor), .groups = "drop")

if (have_votes) {
  d_ch_estV <- st_intersection(precincts, d_ch) |>
    st_make_valid() |>
    st_collection_extract("POLYGON") |>
    dplyr::filter(!st_is_empty(geom)) |>
    mutate(awTot   = totV * as.numeric(st_area(geom) / sarea),
           awMinor = mpV  * as.numeric(st_area(geom) / sarea)) |>
    st_drop_geometry() |>
    group_by(DISTRICT) |>
    summarize(chtotV = sum(awTot), chmpV = sum(awMinor), .groups = "drop")
}
```

```{r map-hull, echo = FALSE, fig.width = map_w, fig.height = map_w * map_asp, fig.cap = paste0("Each district in ", params$state, " with its convex hull.")}
ggplot() +
  geom_sf(data = districts, fill = "#C1E3CD", color = "#006d2c", alpha = 0.8, linewidth = 0.2) +
  geom_sf(data = d_ch, fill = "#F0FFFF", color = "#6EA6C3", alpha = 0.4, linewidth = 0.5) +
  theme_void()
```

## 4.3 Minimum bounding circle composition

The minimum bounding circle is the smallest circle that contains a district. Because it
is a shape operation, we compute it in a projected coordinate system rather than in
geographic coordinates. We pick the NAD83 UTM zone automatically from the data, build
the circle, and transform back to EPSG:4269. We then estimate the minority and vote
share inside each circle in the same way as for the hulls.

```{r mbc, warning = FALSE}
# auto-select the NAD83 UTM zone from the districts' centroid longitude
ctr <- suppressWarnings(st_coordinates(st_centroid(st_union(districts))))
mbc_crs <- 26900L + as.integer(floor((ctr[1] + 180) / 6) + 1)

d_mbc <- districts |>
  dplyr::select(DISTRICT) |>
  st_transform(3395) |>
  lwgeom::st_minimum_bounding_circle() |>
  st_transform(4269) |>
  mutate(mbcArea = st_area(geom))

d_mbc_est <- st_intersection(blockgroups_calc, d_mbc) |>
  st_make_valid() |>
  st_collection_extract("POLYGON") |>
  dplyr::filter(!st_is_empty(geom)) |>
  mutate(awTot   = Total * as.numeric(st_area(geom) / sarea),
         awMinor = minority * as.numeric(st_area(geom) / sarea)) |>
  st_drop_geometry() |>
  group_by(DISTRICT, mbcArea) |>
  summarize(mbcTotal = sum(awTot), mbcMinority = sum(awMinor), .groups = "drop")

if (have_votes) {
  d_mbc_estV <- st_intersection(precincts, d_mbc) |>
    st_make_valid() |>
    st_collection_extract("POLYGON") |>
    dplyr::filter(!st_is_empty(geom)) |>
    mutate(awTot   = totV * as.numeric(st_area(geom) / sarea),
           awMinor = mpV  * as.numeric(st_area(geom) / sarea)) |>
    st_drop_geometry() |>
    group_by(DISTRICT) |>
    summarize(mbctotV = sum(awTot), mbcmpV = sum(awMinor), .groups = "drop")
}
```

```{r map-mbc, echo = FALSE, fig.width = map_w, fig.height = map_w * map_asp, fig.cap = paste0("Each district in ", params$state, " with its minimum bounding circle.")}
ggplot() +
  geom_sf(data = districts, fill = "#C1E3CD", color = "#006d2c", alpha = 0.8, linewidth = 0.2) +
  geom_sf(data = d_mbc, fill = "#F0FFFF", color = "#6EA6C3", alpha = 0.4, linewidth = 0.5) +
  theme_void()
```

## 4.4 Compactness scores and the Cdiff computation

We now compare each district to its compact reference shapes. Three compactness scores
describe how regular a district's shape is. The Polsby-Popper score uses the district
area and perimeter, the convex hull ratio compares the district area to its hull area,
and the Reock score compares the district area to its minimum bounding circle. The
Cdiff scores compare composition instead of shape. `chDiff` is the district's minority
percentage minus the minority percentage of its convex hull, and `mbcDiff` is the same
for the minimum bounding circle. A positive Cdiff suggests packing and a negative Cdiff
suggests cracking. When precinct data is present we compute the same differences for
the minority party vote as `chDiffV` and `mbcDiffV`.

```{r cdiff}
districts_results <- districts_join |>
  left_join(d_ch_est, by = "DISTRICT") |>
  mutate(chPct = round(chMinority / chTotal * 100, 1),
         chDiff = pctMinoritybg - chPct, chAbsDiff = abs(chDiff)) |>
  left_join(d_mbc_est, by = "DISTRICT") |>
  mutate(mbcPct = round(mbcMinority / mbcTotal * 100, 1),
         mbcDiff = pctMinoritybg - mbcPct, mbcAbsDiff = abs(mbcDiff))

if (have_votes) {
  districts_results <- districts_results |>
    left_join(d_ch_estV, by = "DISTRICT") |>
    mutate(chPctV = round(chmpV / chtotV * 100, 1),
           chDiffV = pctMP - chPctV, chAbsDiffV = abs(chDiffV)) |>
    left_join(d_mbc_estV, by = "DISTRICT") |>
    mutate(mbcPctV = round(mbcmpV / mbctotV * 100, 1),
           mbcDiffV = pctMP - mbcPctV, mbcAbsDiffV = abs(mbcDiffV))
}

# Polsby-Popper mixes area and perimeter, so compute it in the projected CRS (meters).
# The hull and circle scores are unit-free ratios, so geographic areas are fine there.
dproj <- st_transform(st_geometry(districts_results), mbc_crs)
districts_results <- districts_results |>
  mutate(darea = st_area(geom),
         compact_shp  = round(as.numeric((4 * pi * st_area(dproj)) /
                                         st_length(st_cast(dproj, "MULTILINESTRING"))^2), 2),
         compact_hull = round(as.numeric(darea / chArea), 2),
         compact_circ = round(as.numeric(darea / mbcArea), 2))
```

Table \@ref(tab:compactness-table) reports the three compactness scores for each district as a
shape reference. They describe only how regular each district's outline is, which is a separate
question from the composition that the Cdiff scores address. A score near 1 is compact and a
score near 0 is stringy.

```{r compactness-table, echo = FALSE}
districts_results |> st_drop_geometry() |>
  dplyr::select(District = DISTRICT,
                `Polsby-Popper` = compact_shp,
                `Convex hull ratio` = compact_hull,
                `Reock` = compact_circ) |>
  arrange(District) |>
  kbl(caption = paste0("Shape-compactness scores by district, ", params$state,
                       " (near 1 is compact, near 0 is stringy).")) |>
  style_tbl()
```

## 4.5 Compact population center composition

We add a third compact reference shape. For each district we find the
population-weighted mean center of its residents and grow a compact district outward
from that center, adding the nearest population until it reaches the target size. The
target size is the state's total population divided by the number of districts, because
redistricting equalizes total population. Centers and distances are computed in a
projected coordinate system so that distances are in meters. As with the other shapes,
composition is measured with the voting-age population.

```{r save-load-race-fragments, warning = FALSE}
race_fragments |> st_write(
  here(output_dir, "race_frags.gpkg"),
  append = FALSE
)

```


```{r cpc-build, warning = FALSE}
n_districts <- dplyr::n_distinct(districts$DISTRICT)
idealpop <- sum(blockgroups_calc$TotalPop) / n_districts

# fragment centroids in the projected CRS (meters)
frag_pts <- race_fragments |> st_centroid() |> st_transform(mbc_crs)
frag_xy  <- st_coordinates(frag_pts)
frag_tab <- st_drop_geometry(frag_pts) |>
  mutate(x = frag_xy[, 1], y = frag_xy[, 2])

# population-weighted mean center of each district
centers <- frag_tab |>
  group_by(DISTRICT) |>
  summarize(cx = sum(x * TotalPop) / sum(TotalPop),
            cy = sum(y * TotalPop) / sum(TotalPop), .groups = "drop") |>
  rename(cpcDistrict = DISTRICT)

# for each center, take the nearest fragments until the target population is reached
cpc_assign <- frag_tab |>
  dplyr::select(frag_id, DISTRICT, x, y, TotalPop, Total, minority, totV, mpV) |>
  cross_join(centers) |>
  mutate(dist = sqrt((x - cx)^2 + (y - cy)^2)) |>
  arrange(cpcDistrict, dist) |>
  group_by(cpcDistrict) |>
  mutate(cumpop = cumsum(TotalPop)) |>
  dplyr::filter(cumpop <= idealpop) |>
  ungroup()
```

Create one spatial layer of locations and their status *vis-a-vis* cpc's.

```{r cpc-fragments, warning = FALSE}
# join cpc assignment table to race_fragments spatial features
cpc_fragments <- race_fragments |>
  left_join(dplyr::select(cpc_assign, frag_id, cpcDistrict), by = "frag_id")

# identity fragments are inside of both the district and it's cpc
cpc_identity <- cpc_fragments |> filter(DISTRICT == cpcDistrict) |> 
  mutate(cpc_type = 'identity')

# captured fragments are inside the district, but not it's cpc, as if the
# compact district boundaries were stretched out to capture these areas
cpc_capture <- race_fragments |> 
  anti_join(st_drop_geometry(cpc_identity), by = "frag_id") |> 
  mutate(cpc_type = 'captured')

# excluded fragments are outside the district, but inside it's cpc, as if
# the compact district boundaries were shrunk / contorted to exclude these areas
cpc_excluded <- cpc_fragments |> 
  filter(DISTRICT != cpcDistrict) |> 
  mutate(cpc_type = 'excluded') 

# count how many cpc's overlapped
cpc_excluded_n <- cpc_excluded |>
  st_drop_geometry() |> 
  group_by(frag_id) |> 
  summarize(excluded_count = n(), .groups ="drop")

# join the count back to the cpc_excluded fragments
cpc_excluded <- cpc_excluded |> 
  left_join(cpc_excluded_n, by = "frag_id")

# merge all three cases of CPCs together and save
cpc_fragments <- bind_rows(cpc_identity, cpc_capture, cpc_excluded)

cpc_fragments |> st_write(here(output_dir, "cpc_fragments.gpkg"),
                          append = FALSE)
```

Summarize CPC results at the district level

```{r cpc-summary, warning = FALSE}
# composition inside each compact population-center district (voting-age)
cpc_est <- cpc_assign |>
  group_by(cpcDistrict) |>
  summarize(cpcTotal = sum(Total), 
            cpcMinority = sum(minority),
            cpcIdealPop = sum(TotalPop),
            cpctotV = sum(totV),
            cpcmpV = sum(mpV)) |> 
  mutate(cpcPct = round(cpcMinority / cpcTotal * 100, 1),
         cpcPctV = round(cpcmpV / cpctotV * 100, 1))

cpc_cases <- cpc_fragments |> 
  group_by(DISTRICT, cpcDistrict, cpc_type) |> 
  summarize(cpcTotal = sum(Total), 
            cpcTargetPop = sum(TotalPop),
            cpcMinority = sum(minority),
            cpctotV = sum(totV),
            cpcmpV = sum(mpV)) |> 
  ungroup()

cpc_cases |> st_write(here(output_dir, "cpc_cases.gpkg"),
                      append = FALSE)

cpcd_identity <- cpc_cases |> 
  filter(cpc_type == 'identity') |> 
  st_drop_geometry() |> 
  mutate(cpcidTotal = cpcTotal, 
         cpcidMinority = cpcMinority,
         cpcidIdealPop = cpcTargetPop,
         cpcidPctMinority = round(cpcidMinority / cpcidTotal * 100, 1),
         cpcidtotV = cpctotV,
         cpcidmpV = cpcmpV,
         cpcidPctV = round(cpcidmpV / cpcidtotV * 100, 1)) |> 
  select(DISTRICT, starts_with("cpcid"))

cpcd_captured <- cpc_cases |> 
  filter(cpc_type == 'captured') |> 
  st_drop_geometry() |> 
  mutate(cpccaTotal = cpcTotal, 
         cpccaMinority = cpcMinority,
         cpccaIdealPop = cpcTargetPop,
         cpccaPctMinority = round(cpccaMinority / cpccaTotal * 100, 1),
         cpccatotV = cpctotV,
         cpccampV = cpcmpV,
         cpccaPctV = round(cpccampV / cpccatotV * 100, 1)) |> 
  select(DISTRICT, starts_with("cpcca"))

cpcd_excluded <- cpc_cases |> 
  filter(cpc_type == 'excluded') |> 
  st_drop_geometry() |> 
  group_by(cpcDistrict) |> 
  summarize(cpcexTotal = sum(cpcTotal),
            cpcexMinority = sum(cpcMinority), 
            cpcexIdealPop = sum(cpcTargetPop),
            cpcextotV = sum(cpctotV),
            cpcexmpV = sum(cpcmpV)) |> 
  mutate(DISTRICT = cpcDistrict,
         cpcexPctMinority = round(cpcexMinority / cpcexTotal * 100, 1),
         cpcexPctV = round(cpcexmpV / cpcextotV * 100, 1)) |> 
  select(DISTRICT, starts_with("cpcex"))

# dissolve the assigned fragments into compact population-center district shapes
cpc_districts <- bind_rows(cpc_excluded, cpc_identity) |>
  group_by(cpcDistrict) |>
  summarize() |> 
  left_join(cpc_est, by="cpcDistrict")

cpc_districts |> st_write(here(output_dir, "cpc_shapes.gpkg"),
                      append = FALSE)

districts_results <- districts_results |>
  left_join(cpc_est, by = c("DISTRICT" = "cpcDistrict")) |>
  left_join(cpcd_identity, by = "DISTRICT") |> 
  left_join(cpcd_captured, by = "DISTRICT") |> 
  left_join(cpcd_excluded, by = "DISTRICT")

districts_results <- districts_results |>
  mutate(cpcDiff = pctMinoritybg - cpcPct, cpcAbsDiff = abs(cpcDiff),
         cpcIdentityPct = cpcidTotal / bgTotal * 100,
         cpcCapturePctDiff = cpccaPctMinority - cpcidPctMinority,
         cpcExcludePctDiff = cpcexPctMinority - cpcidPctMinority,
         cpcDiffV = pctMP - cpcPctV,
         cpcAbsDiffV = abs(cpcDiffV),
         cpcIdentityPctV = cpcidtotV / totV * 100,
         cpcCapturePctDiffV = cpccaPctV - cpcidPctV,
         cpcExcludePctDiffV = cpcexPctV - cpcidPctV,
         )
```


```{r scatterplot_cpcs}
ggplot(data = districts_results, mapping = aes(x = cpcIdentityPct, y = cpcDiff)) + 
  geom_text(aes(label = DISTRICT))
```

```{r cpc-viz}
districts_cpc_viz <- districts_results |> st_drop_geometry() |>
  dplyr::select(DISTRICT, cpcDiff, cpcexTotal, cpcidTotal, cpccaTotal,
                cpcexPctMinority, cpcidPctMinority, cpccaPctMinority,
                cpcIdentityPct) |>
  arrange(desc(cpcDiff), DISTRICT) |>
  mutate(rank = row_number(),
         district_label = paste0("District ", DISTRICT, " (", round(cpcDiff, 1), ")")) |>
  pivot_longer(cols = c(cpcexTotal, cpcidTotal, cpccaTotal),
               names_to = "segment", values_to = "width") |>
  mutate(pct_minority = case_when(
    segment == "cpcexTotal" ~ cpcexPctMinority,
    segment == "cpcidTotal" ~ cpcidPctMinority,
    segment == "cpccaTotal" ~ cpccaPctMinority
  ),
  segment = factor(segment,
                   levels = c("cpcexTotal", "cpcidTotal", "cpccaTotal"))) |>
  group_by(DISTRICT) |>
  arrange(segment, .by_group = TRUE) |>
  mutate(total_width = sum(width),
         xmin = -total_width / 2 + cumsum(width) - width,
         xmax = -total_width / 2 + cumsum(width)) |>
  ungroup()

cpc_viz_extent <- max(districts_cpc_viz$total_width, na.rm = TRUE) / 2

ggplot(districts_cpc_viz,
       aes(y = rank, xmin = xmin, xmax = xmax,
           ymin = rank - 0.35, ymax = rank + 0.35,
           fill = pct_minority)) +
  geom_rect(color = "white", linewidth = 0.4) +
  geom_text(data = dplyr::filter(districts_cpc_viz, segment == "cpcidTotal"),
            aes(x = (xmin + xmax) / 2,
                label = paste0(round(cpcIdentityPct, 1), "%")),
            inherit.aes = FALSE, y = districts_cpc_viz$rank[
              districts_cpc_viz$segment == "cpcidTotal"],
            color = "white", fontface = "bold") +
  scale_y_reverse(breaks = unique(districts_cpc_viz$rank),
                  labels = districts_cpc_viz$district_label[match(
                    unique(districts_cpc_viz$rank), districts_cpc_viz$rank)]) +
  scale_x_continuous(breaks = c(-cpc_viz_extent, 0, cpc_viz_extent),
                     labels = c("excluded", "identity", "captured")) +
  scale_fill_distiller(palette = "YlGnBu", direction = 1,
                       limits = c(0, 100), name = "Minority share (%)") +
  labs(x = NULL,
       y = "District (CPC difference)") +
  theme_bw(base_size = 15) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "bottom")
```
- Y-axis brackets: Delta C
- Y-axis order: Delta C
- X-axis: length proportional to population or votes
- excluded: not inside district, but inside the equal-population circle, i.e. district boundaries were contorted to "exclude" this population from the district, perhaps unfairly
- identity: inside the district AND inside the equal-population circle, i.e. composition of the part of the district that is most compact in shape. Even if composition is unusually high or low, this is a compact shape and likely fair. 
- captured: inside the district, but NOT inside the equal-population circle, i.e. district boundaries were contorted to "capture" this population into the district, perhaps unfairly
- Length of bar segments: total voting-age population
- Color of bar segments: proportion of the segment that is a minority

```{r scatterplot_cpc_votes}
ggplot(data = districts_results, mapping = aes(x = cpcIdentityPctV, y = cpcDiffV)) + 
  geom_text(aes(label = DISTRICT))
```

```{r cpc-viz-votes}
districts_cpc_viz <- districts_results |> st_drop_geometry() |>
  dplyr::select(DISTRICT, cpcDiffV, cpcextotV, cpcidtotV, cpccatotV,
                cpcexPctV, cpcidPctV, cpccaPctV,
                cpcIdentityPctV) |>
  arrange(desc(cpcDiffV), DISTRICT) |>
  mutate(rank = row_number(),
         district_label = paste0("District ", DISTRICT, " (", round(cpcDiffV, 1), ")")) |>
  pivot_longer(cols = c(cpcextotV, cpcidtotV, cpccatotV),
               names_to = "segment", values_to = "width") |>
  mutate(pct_minority = case_when(
    segment == "cpcextotV" ~ cpcexPctV,
    segment == "cpcidtotV" ~ cpcidPctV,
    segment == "cpccatotV" ~ cpccaPctV
  ),
  segment = factor(segment,
                   levels = c("cpcextotV", "cpcidtotV", "cpccatotV"))) |>
  group_by(DISTRICT) |>
  arrange(segment, .by_group = TRUE) |>
  mutate(total_width = sum(width),
         xmin = -total_width / 2 + cumsum(width) - width,
         xmax = -total_width / 2 + cumsum(width)) |>
  ungroup()

cpc_viz_extent <- max(districts_cpc_viz$total_width, na.rm = TRUE) / 2

ggplot(districts_cpc_viz,
       aes(y = rank, xmin = xmin, xmax = xmax,
           ymin = rank - 0.35, ymax = rank + 0.35,
           fill = pct_minority)) +
  geom_rect(color = "white", linewidth = 0.4) +
  geom_text(data = dplyr::filter(districts_cpc_viz, segment == "cpcidTotV"),
            aes(x = (xmin + xmax) / 2,
                label = paste0(round(cpcIdentityPctV, 1), "%")),
            inherit.aes = FALSE, y = districts_cpc_viz$rank[
              districts_cpc_viz$segment == "cpcidTotV"],
            color = "white", fontface = "bold") +
  scale_y_reverse(breaks = unique(districts_cpc_viz$rank),
                  labels = districts_cpc_viz$district_label[match(
                    unique(districts_cpc_viz$rank), districts_cpc_viz$rank)]) +
  scale_x_continuous(breaks = c(-cpc_viz_extent, 0, cpc_viz_extent),
                     labels = c("excluded", "identity", "captured")) +
  scale_fill_distiller(palette = "YlGnBu", direction = 1,
                       limits = c(0, 100), name = "Minority share (%)") +
  labs(x = NULL,
       y = "District (CPC difference)") +
  theme_bw(base_size = 15) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "bottom")
```


```{r save-districts}
districts_results |> st_write(here(output_dir, "district_results.gpkg"),
                              append = FALSE)

districts_results |> 
  st_drop_geometry() |> 
  write_csv(here(output_dir, "district_results.csv"))
```


```{r map-cpc, echo = FALSE, warning = FALSE, fig.width = map_w, fig.height = map_w * map_asp + 0.6, fig.cap = paste0("Compact population-center districts over the actual districts, ", params$state, ".")}
ggplot() +
  geom_sf(data = districts, fill = NA, color = "grey60", linewidth = 0.3) +
  geom_sf(data = cpc_districts, aes(fill = factor(cpcDistrict)), color = NA, alpha = 0.5) +
  labs(fill = "District") +
  theme_void() +
  theme(legend.position = "bottom", plot.margin = margin(0, 0, 0, 0))
```

# 5 Results

## 5.1 Composition difference (Cdiff) table

We report the composition difference two ways, first for the minority party vote and then for the
group population, matching the order of the figures. `r if (have_votes) "Tables \\@ref(tab:results-table-party) and \\@ref(tab:results-table) hold" else "Table \\@ref(tab:results-table) holds"` the
numeric core of the analysis, and every figure below draws from these numbers. Each row is a
district. `pctMinoritybg` is the group's share of the actual district, and `chPct`, `mbcPct`, and
`cpcPct` are the group's share inside the convex hull, the minimum bounding circle, and the compact
population center. `chDiff`, `mbcDiff`, and `cpcDiff` are the actual share minus each of those, so a
positive value is packing and a negative value is cracking. To read it, scan the three Cdiff columns
for each district. A district with large positive values across all three is strongly packed, and
one with large negative values across all three is strongly cracked. Values near zero mean the shape
does not move the composition much. When the three columns disagree in sign, the result depends on
the reference shape, which is itself worth noting. When precinct data is loaded, the partisan table
repeats the same calculation for the minority party vote, with `pctMP` the party's share of the
actual district and `chDiffV`, `mbcDiffV`, and `cpcDiffV` the differences from each reference shape.
Shape compactness is a separate question, reported in Table \@ref(tab:compactness-table). The full
result, including every column, is written to a CSV file.

```{r results-table-party, echo = FALSE, eval = have_votes}
show_cols_v <- c("DISTRICT", "pctMP", "chPctV", "chDiffV", "mbcPctV", "mbcDiffV",
                 "cpcPctV", "cpcDiffV")
tabv <- districts_results |> st_drop_geometry() |>
  dplyr::select(any_of(show_cols_v)) |>
  arrange(DISTRICT)
tabv |>
  kbl(caption = paste0("Cdiff by district for the minority party vote, ",
                         params$state, ".")) |>
  style_tbl() |>
  column_spec(4, bold = TRUE, color = diff_color(tabv$chDiffV)) |>
  column_spec(6, bold = TRUE, color = diff_color(tabv$mbcDiffV)) |>
  column_spec(8, bold = TRUE, color = diff_color(tabv$cpcDiffV)) |>
  add_header_above(ref_header, escape = FALSE) |>
  footnote(general = paste0("Cdiff is the actual district's share minus the reference share. ",
             "Positive suggests <span style='color:#C0405E'>packing</span> and negative ",
             "suggests <span style='color:#3E7EA0'>cracking</span>."),
           general_title = "Note. ", escape = FALSE) |>
  scroll_box(width = "100%")
```

```{r results-table, echo = FALSE}
show_cols <- c("DISTRICT", "pctMinoritybg", "chPct", "chDiff", "mbcPct", "mbcDiff",
               "cpcPct", "cpcDiff")
tab <- districts_results |> st_drop_geometry() |>
  dplyr::select(any_of(show_cols)) |>
  arrange(DISTRICT)
tab |>
  kbl(caption = paste0("Cdiff by district for the ", params$group, " population, ",
                         params$state, ".")) |>
  style_tbl() |>
  column_spec(4, bold = TRUE, color = diff_color(tab$chDiff)) |>
  column_spec(6, bold = TRUE, color = diff_color(tab$mbcDiff)) |>
  column_spec(8, bold = TRUE, color = diff_color(tab$cpcDiff)) |>
  add_header_above(ref_header, escape = FALSE) |>
  footnote(general = paste0("Cdiff is the actual district's share minus the reference share. ",
             "Positive suggests <span style='color:#C0405E'>packing</span> and negative ",
             "suggests <span style='color:#3E7EA0'>cracking</span>."),
           general_title = "Note. ", escape = FALSE) |>
  scroll_box(width = "100%")
```

```{r save-results}
districts_results |> st_drop_geometry() |>
  write_csv(file.path(output_dir, "district_results.csv"))
```

## 5.2 Interpreting the composition difference (Cdiff)

Our approach asks a single question, whether a district's shape changes who it
represents. Rather than judging a district by how irregular it looks, it compares the
composition of the actual district to the composition of a compact reference version of
that same district. The reference can be the convex hull, the minimum bounding circle,
or the compact population center. The metric is the difference between the two
compositions.

$$C_{diff} = C_{district} - C_{reference}$$

Here $C_{district}$ is the minority percentage inside the actual district and
$C_{reference}$ is the minority percentage inside the compact reference shape. `chDiff`,
`mbcDiff`, and `cpcDiff` apply this formula to the convex hull, the minimum bounding
circle, and the compact population center. When precinct data is present, `chDiffV`,
`mbcDiffV`, and `cpcDiffV` apply the same idea to the minority party vote.

There are three cases to read. When $C_{diff}$ is strongly positive, the actual district
holds a higher share of the group than its compact reference, which may suggest packing,
especially where the district is already majority-minority and the group is
over-represented relative to the compact shape. When $C_{diff}$ is strongly negative, the
actual district holds a lower share than its compact reference, which suggests cracking,
dilution, or the exclusion of nearby group members. When $C_{diff}$ is close to zero, the
irregular shape does not meaningfully change the group's composition, so the district may
look non-compact while its shape does not appear to distort racial or partisan
representation.

In the figures that follow, the reading is based on the distance between the actual
district (the black point) and the compact-shape estimates. The black point to the right
of the others points toward packing, and to the left points toward cracking. As you read
the table and figures, consider which districts show the largest gap between the actual
composition and the compact-shape composition, which appear potentially packed, which
appear potentially cracked or diluted, and which stay similar across all geometries.

## 5.3 Composition by district and reference shape

Figure \@ref(fig:figure-composition) has one panel per district. Within a panel, the row labeled Minority Population (%) is the
racial or ethnic composition, and the row labeled Minority Party Vote (%) is the partisan
composition. Minority Population is the share of the voting-age population that belongs to the
group set by `params$group`. Minority Party Vote is the share of the two-party presidential
vote won by the statewide minority party. The vote row appears only when precinct data is
loaded.

Each row carries four points, one for the actual district and one for each compact reference
shape. Black dots are the composition of the actual district. Green dots are the composition
inside its convex hull. Purple dots are the composition inside its minimum bounding circle. Orange
dots are the composition inside its compact population center. The distance from the black dot to
each colored dot is that reference's Cdiff.

To read it, go left to right along each row. When the black point sits to the right of the
colored points, the actual district holds more of the group than its compact references, which
points to packing. When it sits to the left, the district holds less, which points to cracking.
When it sits on top of the colored points, the shape barely changes the composition. When
the colored points fall on both sides of the black point, the three methods disagree, so read that
district with care. The dashed line at 50 percent marks a majority. We do not print the Cdiff
numbers on the plot to keep it readable, so read the exact values from the Cdiff table above.

```{r figure-composition, echo = FALSE, warning = FALSE, fig.height = max(6, 1.5 * n_districts), fig.width = 10, fig.cap = paste0("Composition by district and reference shape, ", params$state, ".")}
combined <- districts_results |> st_drop_geometry() |>
  dplyr::select(DISTRICT, actual = pctMinoritybg, hull = chPct, mbc = mbcPct, cpc = cpcPct) |>
  mutate(type = "Minority Population (%)")
if (have_votes)
  combined <- bind_rows(combined,
    districts_results |> st_drop_geometry() |>
      dplyr::select(DISTRICT, actual = pctMP, hull = chPctV, mbc = mbcPctV, cpc = cpcPctV) |>
      mutate(type = "Minority Party Vote (%)"))
combined <- combined |>
  mutate(DISTRICT = factor(DISTRICT, levels = sort(unique(districts_results$DISTRICT))),
         type = factor(type, levels = c("Minority Population (%)", "Minority Party Vote (%)")))

ggplot(combined, aes(y = type)) + xlim(0, 100) +
  geom_vline(xintercept = 50, color = "grey35", linewidth = 0.7, linetype = "dashed") +
  geom_point(aes(x = hull),   color = "#3E9B63", size = 5) +
  geom_point(aes(x = mbc),    color = "#7D3C98", size = 5) +
  geom_point(aes(x = cpc),    color = "#D98400", size = 5) +
  geom_point(aes(x = actual), color = "#222222", size = 5) +
  facet_wrap(~ DISTRICT, ncol = 1, labeller = label_both) +
  labs(subtitle = paste0(
         "<span style='color:#222222'>Black dots</span>: the actual district. ",
         "<span style='color:#3E9B63'>Green dots</span>: convex-hull reference.<br>",
         "<span style='color:#7D3C98'>Purple dots</span>: minimum-bounding-circle reference. ",
         "<span style='color:#D98400'>Orange dots</span>: compact-population-center reference."),
       x = "Percentage (%)", y = NULL) +
  theme_bw(base_size = 15) +
  theme(plot.subtitle = ggtext::element_markdown())
```

## 5.4 Composition difference (Cdiff) by district

We show the Cdiff two ways, first as bars for the exact size and ranking, then as maps for the
spatial pattern.

### 5.4.1 Cdiff bar chart by district and reference shape

Each bar in `r if (have_votes) "Figures \\@ref(fig:figure-cdiff-bars-party) and \\@ref(fig:figure-cdiff-bars-race)" else "Figure \\@ref(fig:figure-cdiff-bars-race)"` is one district's Cdiff. Bars to the right are packing and bars to the left are
cracking. The three panels are the three reference shapes, in the same order and colors as the
composition figure above. Districts are sorted by convex hull Cdiff, so the most packed sit at
the top and the most cracked at the bottom. `r if (have_votes) "The partisan version comes first and the racial version below it." else "Only the racial version is shown, because no precinct data is loaded."` A packed
district shows a right-pointing bar in every panel, and a cracked district shows a left-pointing
bar in every panel. When a district's bar points right in one panel but left in another, the
reference shapes disagree, which usually means the population is naturally clustered there, so
treat the case with care.

```{r cdiff-bars-setup, include = FALSE}
ord <- districts_results |> st_drop_geometry() |> arrange(chDiff) |> pull(DISTRICT)

make_cdiff_bars <- function(dcols, ttl) {
  tab <- districts_results |> st_drop_geometry() |> dplyr::select(DISTRICT, dplyr::all_of(dcols))
  names(tab)[2:4] <- c("Convex Hull", "Min. Bounding Circle", "Compact Pop. Center")
  tab |>
    pivot_longer(-DISTRICT, names_to = "reference", values_to = "Cdiff") |>
    mutate(DISTRICT = factor(DISTRICT, levels = ord),
           reference = factor(reference, levels = c("Convex Hull", "Min. Bounding Circle", "Compact Pop. Center")),
           direction = ifelse(Cdiff >= 0, "Packing", "Cracking")) |>
    ggplot(aes(x = Cdiff, y = DISTRICT, fill = direction)) +
    geom_col() +
    geom_vline(xintercept = 0, color = "grey35", linewidth = 0.6) +
    scale_fill_manual(values = c("Packing" = "#DC6D85", "Cracking" = "#6EA6C3")) +
    facet_wrap(~ reference, labeller = as_labeller(ref_labels)) +
    labs(x = "Cdiff (percentage points)", y = "District", fill = NULL) +
    theme_bw(base_size = 15) +
    theme(legend.position = "bottom",
          strip.text = ggtext::element_markdown(size = 15, face = "bold"))
}
```

```{r figure-cdiff-bars-party, echo = FALSE, eval = have_votes, fig.height = max(5, 0.9 * n_districts), fig.width = 11, fig.cap = paste0("Cdiff by district for the minority party vote, by reference shape, ", params$state, ".")}
make_cdiff_bars(c("chDiffV", "mbcDiffV", "cpcDiffV"),
                paste0("Cdiff by district, minority party vote (", params$state, ")"))
```

```{r figure-cdiff-bars-race, echo = FALSE, fig.height = max(5, 0.9 * n_districts), fig.width = 11, fig.cap = paste0("Cdiff by district for the ", params$group, " population, by reference shape, ", params$state, ".")}
make_cdiff_bars(c("chDiff", "mbcDiff", "cpcDiff"),
                paste0("Cdiff by district, ", params$group, " population (", params$state, ")"))
```

### 5.4.2 Cdiff choropleth map by district

`r if (have_votes) "Figures \\@ref(fig:figure-cdiff-map-party) and \\@ref(fig:figure-cdiff-map-race)" else "Figure \\@ref(fig:figure-cdiff-map-race)"` draw the same Cdiff on the district map, so the spatial pattern is visible. Each district is
shaded by its Cdiff, red for packed and blue for cracked, with each district number labeled. The
three panels are the three reference shapes, side by side, in the same order as the bar chart. `r if (have_votes) "As with the bars, the partisan map comes first and the racial map below it." else "Only the racial map is shown, because no precinct data is loaded."` To read it, find a district by its number and compare
its color across the three panels. A district that is red in all three panels is robustly packed,
and one that is blue in all three is robustly cracked. A district that changes color between
panels is a case where the choice of reference shape matters.

```{r cdiff-map-setup, include = FALSE}
make_cdiff_map <- function(dcols, ttl) {
  tab <- districts_results |> st_drop_geometry() |> dplyr::select(DISTRICT, dplyr::all_of(dcols))
  names(tab)[2:4] <- c("Convex Hull", "Min. Bounding Circle", "Compact Pop. Center")
  tab <- tab |>
    pivot_longer(-DISTRICT, names_to = "reference", values_to = "Cdiff") |>
    mutate(reference = factor(reference, levels = c("Convex Hull", "Min. Bounding Circle", "Compact Pop. Center")))
  m <- districts |> dplyr::select(DISTRICT) |> left_join(tab, by = "DISTRICT")
  lim <- max(abs(m$Cdiff), na.rm = TRUE)   # symmetric color range
  ggplot(m) +
    geom_sf(aes(fill = Cdiff), color = "grey30", linewidth = 0.2) +
    geom_sf_text(aes(label = DISTRICT), size = 5, fontface = "bold", color = "grey10") +
    scale_fill_gradient2(low = "#6EA6C3", mid = "white", high = "#DC6D85",
                         midpoint = 0, limits = c(-lim, lim),
                         breaks  = c(-lim, 0, lim),
                         labels  = c("Cracking", "0", "Packing"),
                         name = "Cdiff") +
    facet_wrap(~ reference, ncol = 3, labeller = as_labeller(ref_labels)) +
    theme_void(base_size = 18) +
    theme(strip.text = ggtext::element_markdown(size = 20, face = "bold", margin = margin(b = 6)),
          legend.title = element_text(size = 18),
          legend.text = element_text(size = 16),
          legend.key.width = unit(1.4, "cm"))
}
```

```{r figure-cdiff-map-party, echo = FALSE, eval = have_votes, fig.width = 12, fig.height = 4 * map_asp + 0.9, fig.cap = paste0("Cdiff mapped for the minority party vote, by reference shape, ", params$state, ".")}
make_cdiff_map(c("chDiffV", "mbcDiffV", "cpcDiffV"),
               paste0("Cdiff by district, minority party vote (", params$state, ")"))
```

```{r figure-cdiff-map-race, echo = FALSE, fig.width = 12, fig.height = 4 * map_asp + 0.9, fig.cap = paste0("Cdiff mapped for the ", params$group, " population, by reference shape, ", params$state, ".")}
make_cdiff_map(c("chDiff", "mbcDiff", "cpcDiff"),
               paste0("Cdiff by district, ", params$group, " population (", params$state, ")"))
```
