---
title: "`r paste0('A Parsimonious Approach to Gerrymandering Analysis (', params$state, ')')`"
author:
  - "1. Joseph Holler (josephh@middlebury.edu; corresponding author), Middlebury College"
  - "2. Peter Kedron (peterkedron@ucsb.edu), University of California, Santa Barbara"
  - "3. Pyung Kim (pyungkim@ucsb.edu), University of California, Santa Barbara"
  - "4. Emily Zhou (ezhou@ucsb.edu), University of California, Santa Barbara"
date: "`r Sys.Date()`"
knit: (function(input, encoding) { p <- rmarkdown::yaml_front_matter(input)$params; parts <- c(p$state, p$group, p$office, p$focal_party, p$resolution, p$district_year, if (as.character(p$vote_year) != as.character(p$district_year)) p$vote_year, if (nzchar(p$plan)) p$plan); rmarkdown::render(input, encoding = encoding, output_file = paste0("Gerrymander_", gsub("[^A-Za-z0-9]+", "_", paste(parts, collapse = " ")), ".html")) })
output:
  bookdown::html_document2:
    code_folding: hide
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: false
    fig_caption: true
editor_options:
  markdown:
    wrap: sentence
params:
  # Set `state`, `group`, and `district_year`, then Knit. Two steps are not automatic.
  # A free Census API key must be installed once,
  # and the precinct election file must be downloaded by hand.
  # See "How to run this analysis" below.
  state: "Tennessee"          # options. any U.S. state name (e.g. "Tennessee").
                              # the state to analyze.
  group: "Black"              # options. a race, "Hispanic", or "coalition".
                              #   a race is matched by census label, so use "White", "Black",
                              #   "Asian", "American Indian", "Native Hawaiian", or "Some Other Race".
                              #   "Hispanic" is Hispanic of any race;
                              #   "coalition" is everyone except non-Hispanic White alone.
                              # the group to analyze.
  counting: "alone"           # options. "alone" or "combination".
                              #   single race only; ignored for "Hispanic" and "coalition".
                              # "alone" counts that race by itself, on the voting-age population.
                              # "combination" counts it alone or in combination with other races,
                              # from the 2020 DHC file. That count exists only for the total
                              # population, so "combination" reports this group on the
                              # total-population basis (not voting age).
  district_year: 2024         # options. a congressional map year from TIGER (e.g. 2024).
                              # match your research question.
  resolution: "block group"   # options. "block" or "block group".
  vote_year: 2024             # options. the election year of the precinct file
                              # (matches the G<yr> code, e.g. 2024).
  office: "PRE"               # options. "PRE" president, "USS" senate, "USH" house.
                              # the office code in the vote columns.
  focal_party: "auto"         # options. "auto", "DEM", or "REP". the focal party to analyze.
                              # "auto" picks the statewide underdog; "DEM"/"REP" force it.
  plan: ""                    # options. "" or a plan file name in the state folder.
                              # "" uses districts.gpkg or TIGER; a name uses that map.
---

```{=html}
<style>
/* more space before a higher-level heading, less before a lower one, and tight
   spacing inside a section so one numbered part reads as a single block */
h1 { margin-top: 3.4em; margin-bottom: 0.4em; }
h2 { margin-top: 2.2em; margin-bottom: 0.3em; }
h3 { margin-top: 1.3em; margin-bottom: 0.3em; }
h4 { margin-top: 0.8em; margin-bottom: 0.2em; }
p  { margin-top: 0.2em; margin-bottom: 0.5em; }
.figure, img { margin-top: 0.4em; margin-bottom: 0.4em; }

/* figure captions ("Figure N: ...") large and dark so they read as titles */
.figure .caption, div.figure p.caption, p.caption, .caption {
  font-size: 1.5rem !important;
  font-weight: 700 !important;
  color: #16232a !important;
  margin-top: 0.6em;
  text-align: left;
  line-height: 1.35;
}
/* keep the table body compact without shrinking the caption (font_size in
   kable_styling would force the caption to a fixed size and block styling) */
.table, table.table { font-size: 14px; }
/* table captions sit above the table, prominent rather than faint grey.
   rem keeps them root-relative so the 14px table body does not shrink them */
table > caption, .table > caption, table caption, caption {
  caption-side: top !important;
  font-size: 1.5rem !important;
  font-weight: 700 !important;
  color: #16232a !important;
  text-align: left !important;
  padding-bottom: 0.4em;
  line-height: 1.35;
}
/* the "Table N:" label bold (kableExtra wraps it in a span carrying the tab id) */
caption span[id^="tab:"] { font-weight: 700; }
</style>
```

# Abstract {.unnumbered}

This is a study of gerrymandering in the United States. We introduce a parsimonious
composite approach that compares each district's racial and partisan composition to
that of compact reference shapes. We implement the entire workflow so that it can be
applied to any U.S. state from standard, publicly available data.

## Study Metadata {.unnumbered}

- `Key words`: gerrymandering, compactness, convex hull, racial and partisan representation
- `Subject`: Social and Behavioral Sciences: Geography: Geographic Information Sciences
- `Date created`: 2025-02-17
- `Date modified`: `r Sys.Date()`
- `Spatial Coverage`: United States; one state per run, selected via `params$state` (this run: `r params$state`)
- `Spatial Resolution`: Census block groups (demographics) and voting precincts (elections)
- `Spatial Reference System`: EPSG:4269 NAD 1983 Geographic Coordinate System
- `Temporal Coverage`: 2020 decennial census population; 2024 presidential election results
- `Temporal Resolution`: Decennial census

# 1 Study design

## 1.1 Overview

This is an original, exploratory study based on the literature on gerrymandering
metrics. We evaluate the usefulness of a new gerrymandering metric that measures how
much a congressional district's representativeness differs from that of compact
reference shapes derived from it. These reference shapes are its convex hull, its
minimum bounding circle, and a compact district grown from its population-weighted
mean center.

Our contribution is methodological and geographic rather than a claim about political
science theory. Most geometric measures judge a district by the compactness of its shape
alone. We instead introduce a composition difference (ΔC) that couples a district's racial
or partisan composition with the geometry of compact reference shapes, and we implement it
as a reproducible workflow that runs for any U.S. state from standard public data. Framing
gerrymandering this way connects it to long-standing geographic concerns with the
modifiable areal unit problem, scale, and edge effects.

## 1.2 Bias and threats to validity

This study is explicitly an investigation into the modifiable areal unit problem.
Shape-based measures are sensitive to edge effects and scale, because complex borders
formed by natural features such as coastlines and rivers vary greatly in perimeter
depending on the scale of analysis. We hope this study establishes a method that is
more robust to the threats to validity caused by scale and edge effects in studies of
gerrymandering and district shape.

# 2 How to run this analysis

1. Set the analysis in the header. `params$state` is the state, `params$group` is the
   racial or ethnic group to analyze (a race such as "Black" or "Asian", or "Hispanic"),
   and `params$district_year` is the congressional map year.
2. Choose `params$district_year` for your research purpose. To evaluate a particular
   election, use the year of the map that election was run under, since some states redrew
   their map mid-decade. You can also override TIGER entirely by placing a specific plan as
   `districts.gpkg` in the state folder `data/raw/private/<state>/`.
3. Get a free Census API key at https://api.census.gov/data/key_signup.html and
   install it once by running `census_api_key("YOUR KEY", install = TRUE)` in the
   RStudio console, then restart R. This lets the document download census block
   groups and congressional districts. You can skip it if you supply those as local
   files.
4. To include the voting layer, download that state's "2024 General Election Precinct
   Boundaries and Election Results" from the Redistricting Data Hub and unzip it into
   `data/raw/private/<state>/`. Redistricting Data Hub data cannot be redistributed, so keep
   the `private` folder out of the shared repo. The code finds the `_all_` precinct shapefile
   inside that folder automatically, so you can leave the downloaded folder as is. The
   congressional districts and census block groups download automatically, so you do not need those.
5. Knit the document. The code reads `params$state` and loads that state's data.

# 3 Materials and procedure

## 3.1 Computational environment

We use `tidyverse`, `sf`, and `lwgeom` for data wrangling and spatial operations. We
use `tidycensus` and `tigris` to obtain census demographics and district geographies. We
use `ggplot2` (part of `tidyverse`) for mapping and `knitr` with `kableExtra` for tables.
The setup chunk installs any missing package so the analysis runs on a clean machine.

```{r environment-setup, include = FALSE}
options(scipen = 999, digits = 3)

# packages used in this analysis
packages <- c("tidyverse", "here", "sf", "lwgeom", "geosphere",
              "tidycensus", "tigris", "knitr", "kableExtra", "ggtext", "patchwork", "bookdown")

# install any that are missing, then load them
for (p in packages) {
  if (!requireNamespace(p, quietly = TRUE))
    install.packages(p, repos = "https://cloud.r-project.org")
  suppressPackageStartupMessages(library(p, character.only = TRUE))
}

# use the GEOS planar engine rather than the s2 spherical engine. Precinct files
# from different states vary in quality, and s2 halts on tiny geometry defects
# such as duplicate vertices. GEOS repairs and measures them without stopping.
sf::sf_use_s2(FALSE)
```

```{r rec-environment, warning = FALSE, message = FALSE}
# default knitr options
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, fig.width = 8,
                      warning = FALSE, message = FALSE)

# one consistent table style used throughout, with a colored bold header row
style_tbl <- function(kb) {
  kb |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width = TRUE, position = "center") |>
    kableExtra::row_spec(0, bold = TRUE, color = "white", background = "#3B6B7A")
}

# color for Delta-C cells, matching the maps and bars: concentration (positive) red,
# dilution (negative) blue, near-zero grey
diff_color <- function(x) ifelse(is.na(x), "#555555",
                          ifelse(x > 0, "#C0405E", ifelse(x < 0, "#3E7EA0", "#555555")))

# one color per reference shape, used to tint the facet strip labels (figures)
# and the grouped column headers (tables): hull green, circle pink, center orange
ref_col <- c("Convex Hull" = "#3E9B63", "Min. Bounding Circle" = "#7D3C98",
             "Compact Pop. Center" = "#D98400")
ref_labels <- setNames(
  paste0("<span style='color:", ref_col, "'>", names(ref_col), "</span>"),
  names(ref_col))

# grouped column header for the results tables (plain, no color)
ref_header <- c(" " = 2, "Convex hull" = 2, "Min. bounding circle" = 2,
                "Compact pop. center" = 2)

# optionally record the R session and package citations; never halt the knit if
# the folders are missing or not writable
try(silent = TRUE, {
  dir.create(here("procedure", "environment"), recursive = TRUE, showWarnings = FALSE)
  writeLines(capture.output(sessionInfo()),
             here("procedure", "environment", paste0("r-environment-", Sys.Date(), ".txt")))
  knitr::write_bib(c(packages, "base"), file = here("software.bib"))
})
```

## 3.2 Data and variables

### 3.2.1 Data sources

The workflow reads each dataset from a local file when present, and automatically
downloads the ones that have a programmatic source for `params$state`. Only the
precinct data must be downloaded by hand.

- **Congressional districts.** The boundaries of the districting plan under study.
  Downloaded automatically from the U.S. Census TIGER files with the `tigris`
  package, or read from a local plan file obtained from the state legislature or the
  Redistricting Data Hub.
- **Voting precincts with election results.** Precinct boundaries joined to
  general-election returns, used for the partisan analysis. There is no public API,
  so this file is downloaded by hand from the Redistricting Data Hub at
  https://redistrictingdatahub.org (free account). On the state page, choose the
  dataset named "General Election [State] VTD (Precinct) Level Results and Boundaries"
  for the year of interest (for example 2024) and download the shapefile. Vote
  columns follow the `G<yr>PRE<party><cand>` naming convention, for example
  `G24PRERTRU` for 2024 presidential votes for the Republican candidate.
- **Census block groups.** Population by race and Hispanic ethnicity, from the 2020
  Census PL 94-171 redistricting file, downloaded automatically with the `tidycensus`
  package: table `P3` (race, 18 and over), `P4` (Hispanic origin by race, 18 and over),
  `P2` (Hispanic origin by race, total population), and `P1` (total population). Block
  group geometry is included in the same download.
- **Race alone or in combination (only when `params$counting` is "combination").** The
  count of the group counted alone or in combination with one or more other races, read
  from the 2020 Census Demographic and Housing Characteristics (DHC) file, table `P7`,
  downloaded automatically with `tidycensus`. This is needed because the PL 94-171 file
  above reports only race-alone counts, so an inclusive count has no source there. The
  DHC publishes it for the total population and not for voting age, which is why the
  "combination" option reports the group on the total-population basis.

The `P3` variable metadata that labels which columns count as the minority group is a
supporting lookup. It is downloaded automatically with `tidycensus::load_variables`
and cached as `census2020pl_vars.csv`.

This analysis uses precinct data from the Redistricting Data Hub, which asks that products
built on their data carry the attribution "This map was created using data from the
Redistricting Data Hub, redistrictingdatahub.org."

### 3.2.2 Loading the data

We set the file paths from `params$state` and load each dataset, downloading the ones
with a programmatic source when a local copy is not present.

```{r load-data, results = "hide"}
# per-state data folder and file paths (edit names to match your files)
state_slug      <- gsub("[^A-Za-z0-9]+", "_", params$state)
# all inputs live under private: RDH election data cannot be redistributed and block data
# is too large to commit, so nothing here is committed to the shared repo
data_dir        <- here("data", "raw", "private", state_slug)
geo             <- params$resolution                            # "block" or "block group"

# result-set label encodes the analysis settings so 2+ runs of the same state do not
# overwrite each other; it names the output folder and is recorded alongside the outputs
# encode every setting that changes the output, in the same order as the HTML file name,
# so two runs that differ in any of them do not overwrite each other
result_parts <- c(params$state, params$group, params$office, params$focal_party,
                  params$resolution, params$district_year)
if (as.character(params$vote_year) != as.character(params$district_year))
  result_parts <- c(result_parts, params$vote_year)
if (nzchar(params$plan)) result_parts <- c(result_parts, params$plan)
result_set   <- gsub("[^A-Za-z0-9]+", "_", paste(result_parts, collapse = " "))
output_dir   <- here("data", "derived", "private", result_set)

# districts: set params$plan to a file name in the state folder to analyze that specific map,
# otherwise use districts.gpkg (if present) or the TIGER map for params$district_year
districts_file  <- if (nzchar(params$plan)) file.path(data_dir, params$plan) else file.path(data_dir, "districts.gpkg")
blockgroup_file <- file.path(data_dir, if (geo == "block") "block.gpkg" else "block_groups.gpkg")
census_metadata_file <- here("data", "metadata", "census2020pl_vars.csv")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

# Tables and figures are written to the shared data/output folder (data itself goes to
# the derived result folder above). Figures are named "Figure N <title>"; the number is a
# running count that follows the order figures appear, matching the report numbering.
output_report_dir <- here("data", "output")
dir.create(output_report_dir, recursive = TRUE, showWarnings = FALSE)
# clear stale "Figure N ..."/"Table N ..." files from an earlier run so a run that produces
# fewer or differently titled outputs does not leave orphans (only our own naming pattern
# is matched, so nothing else in the folder is touched)
unlink(list.files(output_report_dir,
                  pattern = "^(Figure|Table) [0-9]+ .*\\.(png|csv)$", full.names = TRUE))
knitr::opts_chunk$set(fig.path = file.path(output_report_dir, ""))

# make a caption filesystem-safe (drop punctuation such as the ΔC glyph and colons)
clean_title <- function(x) {
  x <- gsub("[^A-Za-z0-9 ]+", " ", x)
  substr(gsub("\\s+", " ", trimws(x)), 1, 90)
}

# rename each rendered figure to "Figure N <title>.png" as knitr writes it
.fig_counter <- 0L
knitr::opts_chunk$set(fig.process = function(path, options) {
  .fig_counter <<- .fig_counter + 1L
  cap <- options$fig.cap
  if (is.null(cap) || !nzchar(cap)) cap <- options$label
  dest <- file.path(output_report_dir, sprintf("Figure %d %s.png", .fig_counter, clean_title(cap)))
  ok <- suppressWarnings(try(file.rename(path, dest), silent = TRUE))
  if (inherits(ok, "try-error") || !isTRUE(ok)) return(path)
  dest
})

# write each table to data/output as "Table N <title>.csv", counted in appearance order
.tab_counter <- 0L
save_output_table <- function(df, title) {
  .tab_counter <<- .tab_counter + 1L
  f <- file.path(output_report_dir, sprintf("Table %d %s.csv", .tab_counter, clean_title(title)))
  try(readr::write_csv(as.data.frame(df), f), silent = TRUE)
  invisible(df)
}

# record the run settings (a copy of the header params) inside the result folder
writeLines(vapply(names(params), function(k) paste0(k, ": ", params[[k]]), character(1)),
           file.path(output_dir, "run_params.txt"))

# precincts: use the Redistricting Data Hub "_all_" precinct shapefile found anywhere
# under the state folder, or a file named precincts.shp; fall back if none is present
precincts_file  <- {
  hits <- list.files(data_dir, pattern = "_all_prec\\.shp$", recursive = TRUE, full.names = TRUE)
  if (!length(hits)) hits <- list.files(data_dir, pattern = "precincts\\.shp$", recursive = TRUE, full.names = TRUE)
  if (length(hits)) hits[1] else file.path(data_dir, "precincts.shp")
}

# make sure the folders used to cache downloaded data exist
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(census_metadata_file), recursive = TRUE, showWarnings = FALSE)

# districts: a local plan file if present, otherwise the enacted districts from TIGER
if (file.exists(districts_file)) {
  districts <- st_read(districts_file)
} else {
  districts <- tigris::congressional_districts(state = params$state, cb = TRUE,
                                               year = params$district_year)
}

# precincts with election results: local file only (manual download), optional
have_votes <- file.exists(precincts_file)
if (have_votes) precincts <- st_read(precincts_file)

# census variable metadata: cached csv, or query and cache
if (file.exists(census_metadata_file)) {
  census2020pl_vars <- read.csv(census_metadata_file)
} else {
  census2020pl_vars <- load_variables(2020, "pl")
  write.csv(census2020pl_vars, census_metadata_file)
}

# census block groups: P3 (race for 18+), P4 (Hispanic ethnicity for 18+, voting age),
# P2 (Hispanic ethnicity for all ages, total population), and P1_001N (total population
# all ages); cached, or queried and cached
if (file.exists(blockgroup_file)) {
  blockgroups <- st_read(blockgroup_file)
} else {
  blockgroups <- get_decennial(geography = geo, sumfile = "pl", table = "P3",
                               year = 2020, state = params$state, output = "wide",
                               geometry = TRUE, keep_geo_vars = TRUE)
  # Hispanic or Latino ethnicity by race: voting age (table P4) and all ages (table P2).
  # P2 mirrors P4 category by category, so it gives the same groups on the total population.
  hisp     <- get_decennial(geography = geo, sumfile = "pl", table = "P4",
                            year = 2020, state = params$state, output = "wide")
  hisp_tot <- get_decennial(geography = geo, sumfile = "pl", table = "P2",
                            year = 2020, state = params$state, output = "wide")
  # total population of all ages, used later to size the compact districts
  totalpop <- get_decennial(geography = geo, sumfile = "pl",
                            variables = "P1_001N", year = 2020, state = params$state,
                            output = "wide")
  blockgroups <- blockgroups |>
    left_join(dplyr::select(hisp, GEOID, dplyr::starts_with("P4_")), by = "GEOID") |>
    left_join(dplyr::select(hisp_tot, GEOID, dplyr::starts_with("P2_")), by = "GEOID") |>
    left_join(dplyr::select(totalpop, GEOID, P1_001N), by = "GEOID")
  st_write(blockgroups, blockgroup_file)
}

# an older cached file may hold only some P4 columns; ensure the full P4 table is present
needed_p4 <- c("P4_001N", "P4_002N", "P4_005N", "P4_006N", "P4_007N",
               "P4_008N", "P4_009N", "P4_010N", "P4_011N")
if (!all(needed_p4 %in% names(blockgroups))) {
  hisp <- get_decennial(geography = geo, sumfile = "pl", table = "P4",
                        year = 2020, state = params$state, output = "wide")
  blockgroups <- blockgroups |>
    dplyr::select(-dplyr::any_of(grep("^P4_", names(blockgroups), value = TRUE))) |>
    left_join(dplyr::select(hisp, GEOID, dplyr::starts_with("P4_")), by = "GEOID")
}

# likewise ensure the full P2 (total-population) table is present in an older cache
needed_p2 <- c("P2_001N", "P2_002N", "P2_005N", "P2_006N", "P2_007N",
               "P2_008N", "P2_009N", "P2_010N", "P2_011N")
if (!all(needed_p2 %in% names(blockgroups))) {
  hisp_tot <- get_decennial(geography = geo, sumfile = "pl", table = "P2",
                            year = 2020, state = params$state, output = "wide")
  blockgroups <- blockgroups |>
    dplyr::select(-dplyr::any_of(grep("^P2_", names(blockgroups), value = TRUE))) |>
    left_join(dplyr::select(hisp_tot, GEOID, dplyr::starts_with("P2_")), by = "GEOID")
}
```

## 3.3 Data processing

### 3.3.1 Spatial transformations

The districts, precincts, and block groups arrive in different coordinate reference
systems. To perform spatial operations we transform them all to a common CRS, the
NAD 1983 geographic coordinate system (EPSG:4269). We also standardize the district
identifier column, whose name varies by data source, to `DISTRICT`.

```{r transform}
districts <- st_make_valid(st_transform(districts, 4269))
names(districts)[names(districts) == attr(districts, "sf_column")] <- "geom"
st_geometry(districts) <- "geom"

# standardize the district id column to "DISTRICT" (names vary by source)
dcol <- c("DISTRICT", "DISTRICT_I", grep("^CD[0-9]+FP$", names(districts), value = TRUE))
dcol <- dcol[dcol %in% names(districts)][1]
districts$DISTRICT <- as.integer(as.character(districts[[dcol]]))

blockgroups <- st_make_valid(st_transform(blockgroups, 4269))
names(blockgroups)[names(blockgroups) == attr(blockgroups, "sf_column")] <- "geom"
st_geometry(blockgroups) <- "geom"

if (have_votes) {
  precincts <- st_make_valid(precincts) |> st_transform(4269)
  names(precincts)[names(precincts) == attr(precincts, "sf_column")] <- "geom"
  st_geometry(precincts) <- "geom"
}

# aspect ratio (height / width) of any layer's bounding box, used to size maps so they do
# not leave whitespace. Reference shapes such as bounding circles extend past the state, so
# each of those maps is sized to its own extent rather than the state's.
bbox_asp <- function(x) {
  b <- st_bbox(x)
  as.numeric((b["ymax"] - b["ymin"]) /
             ((b["xmax"] - b["xmin"]) * cospi(mean(c(b["ymin"], b["ymax"])) / 180)))
}
map_w   <- 8   # map width in inches
map_asp <- bbox_asp(districts)
```


```{r map-districts, echo = FALSE, fig.width = map_w, fig.height = map_w * map_asp, fig.cap = paste0("Congressional districts of ", params$state, ", each labeled by its congressional district number.")}
ggplot(districts) +
  geom_sf(fill = "#C1E3CD", color = "#006d2c", alpha = 0.5) +
  geom_sf_text(aes(label = DISTRICT), color = "black", size = 5, fontface = "bold") +
  theme_void()
```

### 3.3.2 Compute variables of interest

We compute the percentage of the population 18 years and older that belongs to the group
named by `params$group`, using table P4, which breaks race out on the not-Hispanic side. For a
race such as Black or Asian, `params$counting` chooses whether to count people who report that
race alone (exclusive) or alone or in combination with another race (inclusive). The
alone-or-in-combination count is published only for the total population (from the 2020 DHC
file), so choosing `combination` reports that group on the total population rather than the
voting-age population. For Hispanic or Latino we use the Hispanic total of any race, and for
`coalition` we count everyone except non-Hispanic White alone.

```{r minority-vars}
group_label <- params$group           # a race (Black, Asian, ...), "Hispanic", or "coalition"
counting <- tolower(params$counting)   # "alone" (exclusive) or "combination" (inclusive)
total_basis <- FALSE                   # combination reports the group on the total population

if (tolower(group_label) == "coalition") {
  # all minorities: everyone except non-Hispanic White alone (P4_001N - P4_005N),
  # written as the sum of the non-White P4 categories
  minority_vars <- census2020pl_vars |>
    dplyr::filter(name %in% c("P4_002N", "P4_006N", "P4_007N", "P4_008N",
                              "P4_009N", "P4_010N", "P4_011N")) |>
    dplyr::select(name, label)
} else if (tolower(group_label) == "hispanic") {
  # Hispanic or Latino of any race, population 18 and over (P4_002N)
  minority_vars <- census2020pl_vars |>
    dplyr::filter(name == "P4_002N") |>
    dplyr::select(name, label)
} else if (counting == "combination") {
  # alone-or-in-combination is published only for the TOTAL population (2020 DHC tables
  # P6/P7), not for voting age, so "combination" switches this group to the total-population
  # basis. We use the not-Hispanic "alone or in combination" count (table P7) for the group,
  # fetched from DHC and joined to the block groups.
  aoic_var <- NA_character_
  try({
    dhc_vars <- load_variables(2020, "dhc")
    aoic <- dhc_vars |>
      dplyr::filter(str_detect(tolower(label), "alone or in combination"),
                    str_detect(label, "Not Hispanic"),
                    str_detect(label, group_label))
    aoic_var <- aoic$name[1]
    if (!is.na(aoic_var) && !aoic_var %in% names(blockgroups)) {
      aoic_bg <- get_decennial(geography = geo, sumfile = "dhc", variables = aoic_var,
                               year = 2020, state = params$state, output = "wide")
      blockgroups <- blockgroups |>
        left_join(dplyr::select(aoic_bg, GEOID, dplyr::all_of(aoic_var)), by = "GEOID")
    }
  }, silent = TRUE)
  if (!is.na(aoic_var) && aoic_var %in% names(blockgroups)) {
    minority_vars <- data.frame(name = aoic_var, label = aoic$label[1],
                                stringsAsFactors = FALSE)
    total_basis <- TRUE
  } else {
    warning("DHC alone-or-in-combination is not available for ", group_label,
            "; falling back to 'alone' on the voting-age population.")
    minority_vars <- census2020pl_vars |>
      dplyr::filter(str_detect(name, "^P4"), str_detect(label, group_label),
                    str_detect(label, "alone")) |>
      dplyr::select(name, label)
  }
} else {
  # a single race counted alone, from the not-Hispanic P4 categories (voting age)
  minority_vars <- census2020pl_vars |>
    dplyr::filter(str_detect(name, "^P4"), str_detect(label, group_label),
                  str_detect(label, "alone")) |>
    dplyr::select(name, label)
}

# the same group on the total population of all ages, using the parallel P2 codes.
# P2 mirrors P4 category by category (P4_006N is Black alone at 18+, P2_006N is Black alone
# at all ages). The sub() is a no-op when combination already uses a total-population variable.
minority_vars_tot <- minority_vars |>
  dplyr::mutate(name = sub("^P4_", "P2_", name))

# the population basis this run reports the group on, named in captions and prose below
basis_word <- if (total_basis) "total population" else "voting-age population"
```

```{r table1, echo = FALSE}
save_output_table(minority_vars, "Selected census variables for the group of interest")
minority_vars |>
  rename(Variable = name, Label = label) |>
  kbl(caption = "Selected census variables for the group of interest.") |>
  style_tbl() |>
  scroll_box(height = "300px")
```

For each block group we sum the selected columns into `minority`, use the matching population
total as `pop_basis` (the voting-age `P3_001N`, or the total population `P1_001N` when counting is
combination), and compute `pctMinority`. We keep the all-ages total as `pop_total`, and compute
the area of each block group for later area-weighted reaggregation.

```{r bg-vars}
blockgroups_calc <- blockgroups |>
  dplyr::filter(!st_is_empty(geom)) |>
  rowwise() |>
  mutate(minority = sum(c_across(all_of(minority_vars$name))),         # numerator on the run basis
         minority_tot = sum(c_across(all_of(minority_vars_tot$name)))) |>  # all ages (P2)
  ungroup() |>
  mutate(sarea = st_area(geom),
         # pop_basis is the analysis denominator on the run's basis: voting-age (P3_001N)
         # normally, or total population (P1_001N) when counting is "combination", so the
         # numerator and denominator share one basis. pop_total is always the all-ages total.
         pop_basis = if (total_basis) P1_001N else P3_001N,
         pop_total = P1_001N,  # all ages, used to size the compact districts and total-pop composition
         pctMinority = minority / pop_basis * 100) |>
  dplyr::select(GEOID, sarea, minority, minority_tot, pop_basis, pop_total, pctMinority)
```

For the precincts we compute `totV`, the two-party total for the election named by
`params$office` and `params$vote_year`, and `mpV`, the focal party vote. The focal
party is the statewide underdog by default, the party with fewer total votes, or the
party forced by `params$focal_party`.

```{r precinct-vars}
fp_disp <- "focal party"   # default caption label; replaced with the party name when votes load
if (have_votes) {
  # vote columns follow the RDH code G<yr><office><party>..., e.g. G24PRER for a 2024
  # presidential Republican. We pick them by position from the params rather than a fixed
  # string: year is characters 2-3, office is 4-6, party is the 7th character (R or D).
  yr2 <- substr(as.character(params$vote_year), 3, 4)
  nm  <- names(precincts)
  sel <- substr(nm, 1, 1) == "G" &
         substr(nm, 2, 3) == yr2 &
         substr(nm, 4, 6) == toupper(params$office)
  rep_cols <- nm[sel & substr(nm, 7, 7) == "R"]
  dem_cols <- nm[sel & substr(nm, 7, 7) == "D"]

  # if the chosen office is not present in this precinct file, skip the partisan analysis
  # rather than build empty vote columns (e.g. House races are coded per district, not as a
  # single statewide R/D pair)
  if (length(rep_cols) == 0 || length(dem_cols) == 0) {
    warning("No vote columns match office = ", params$office,
            "; skipping the partisan analysis for this run.")
    have_votes <- FALSE
  }
}

if (have_votes) {
  # focal party: forced by params$focal_party, or the statewide underdog when "auto"
  Rtot <- sum(rowSums(st_drop_geometry(precincts)[rep_cols]), na.rm = TRUE)
  Dtot <- sum(rowSums(st_drop_geometry(precincts)[dem_cols]), na.rm = TRUE)
  focal_party <- if (toupper(params$focal_party) == "DEM") "Democratic"
                 else if (toupper(params$focal_party) == "REP") "Republican"
                 else if (Dtot <= Rtot) "Democratic" else "Republican"
  mp_cols <- if (focal_party == "Democratic") dem_cols else rep_cols

  # label for figure captions and legends, naming the actual party, e.g. "focal party (Democratic)"
  fp_disp <- paste0("focal party (", focal_party, ")")

  precincts <- precincts |>
    mutate(totV = rowSums(across(all_of(c(rep_cols, dem_cols)))),
           mpV  = rowSums(across(all_of(mp_cols))),
           sarea = st_area(geom))
}
```


```{r disaggregate-votes}
if (have_votes) {
  # Allocate each precinct's votes to intersecting block groups by voting-age population.
  vote_fragments <- st_intersection(
    precincts |>
      mutate(precinct_id = dplyr::row_number()) |>
      dplyr::select(precinct_id, totV, mpV),
    blockgroups_calc |>
      dplyr::select(GEOID, pop_basis, sarea)
  ) |>
    mutate(bg_weight = pop_basis * as.numeric(st_area(geom) / sarea),
           prec_totV = totV,
           prec_mpV = mpV) |>
    group_by(precinct_id) |>
    mutate(weight = bg_weight / sum(bg_weight),
           totV = prec_totV * weight,
           mpV = prec_mpV * weight) |>
    ungroup() |>
    group_by(GEOID) |>
    summarize(totV = sum(totV), mpV = sum(mpV), .groups = "drop")

  blockgroups_calc <- blockgroups_calc |>
    left_join(st_drop_geometry(vote_fragments), by = "GEOID") |>
    mutate(totV = coalesce(totV, 0),
           mpV = coalesce(mpV, 0))
}

# ensure vote columns exist even without precincts, so downstream code does not break
if (!have_votes) {
  blockgroups_calc$totV <- 0
  blockgroups_calc$mpV  <- 0
}
```

`r if (have_votes) paste0("For this run the focal party is the ", focal_party, " party, based on the ", params$vote_year, " ", params$office, " election.") else "The partisan analysis is skipped for this run, because no precinct file is present or the chosen office is not in that file."`

### 3.3.3 Descriptive statistics (racial and ethnic composition by district)

Table \@ref(tab:desc-comp) summarizes the racial and ethnic composition of the voting-age
population before the analysis. The first row is the statewide baseline, and each following row is a district.
District shares come from the same area-weighted reaggregation of the block groups used in
the rest of the analysis. The categories come from table P4, which cross-tabs race and Hispanic
ethnicity into mutually exclusive groups (non-Hispanic by race counted alone, plus Hispanic of any
race), so they do not double-count and add up to about 100 percent. The Coalition column is everyone
except non-Hispanic White alone. To read it, compare each district row to the Statewide row on top. A district
whose share is far above the statewide share holds a lot of that group, and one far below holds
little. Big swings between districts are the raw material the rest of the analysis examines,
since gerrymandering works by concentrating a group in a few districts (packing) or spreading it
thin across many (cracking).

```{r awr-functions, include = FALSE}
# Reusable area-weighted reaggregation in two directional steps, defined here because
# both the descriptive composition table below and the analysis in section 4 use them.
# F1 disaggregate: cut source units by target polygons and area-weight the value columns
#    into fragments. F2 aggregate: sum those fragment values up into each target unit.
disaggregate_to_fragments <- function(source, target, value_cols) {
  source |>
    mutate(src_area_disagg = st_area(geom)) |>
    st_intersection(target) |>
    st_make_valid() |>
    st_collection_extract("POLYGON") |>   # drop line/point slivers from planar overlaps
    dplyr::filter(!st_is_empty(geom)) |>
    st_cast("MULTIPOLYGON") |> st_cast("POLYGON") |>
    mutate(frag_w_disagg = as.numeric(st_area(geom) / src_area_disagg)) |>
    mutate(dplyr::across(dplyr::all_of(value_cols), ~ .x * frag_w_disagg),
           frag_id = dplyr::row_number()) |>
    dplyr::select(-src_area_disagg, -frag_w_disagg)
}
aggregate_to_target <- function(fragments, id_cols, value_cols) {
  fragments |>
    st_drop_geometry() |>
    group_by(dplyr::across(dplyr::all_of(id_cols))) |>
    summarize(dplyr::across(dplyr::all_of(value_cols), sum), .groups = "drop")
}
```

```{r desc-comp, echo = FALSE, warning = FALSE, message = FALSE}
# mutually exclusive P4 categories (voting age): non-Hispanic by race counted alone,
# plus Hispanic of any race. They do not overlap and sum to the total.
p4_vars <- c(`White (NH)` = "P4_005N", `Black (NH)` = "P4_006N",
             `AIAN (NH)` = "P4_007N", `Asian (NH)` = "P4_008N",
             `NHPI (NH)` = "P4_009N", `Other (NH)` = "P4_010N",
             `Two+ (NH)` = "P4_011N", `Hispanic` = "P4_002N")

# per-district composition by the same area-weighted reaggregation used in the analysis
# (disaggregate_to_fragments/aggregate_to_target), so the descriptive table and the
# results share one intersection routine
bg_comp <- blockgroups |>
  dplyr::filter(!st_is_empty(geom)) |>
  dplyr::select(pop_vap = P4_001N, dplyr::all_of(unname(p4_vars)))

dist_comp <- disaggregate_to_fragments(bg_comp, districts,
                                       c("pop_vap", unname(p4_vars))) |>
  aggregate_to_target("DISTRICT", c("pop_vap", unname(p4_vars))) |>
  arrange(DISTRICT)

# statewide totals
sw <- blockgroups |> st_drop_geometry() |>
  dplyr::summarize(pop_vap = sum(P4_001N),
                   dplyr::across(dplyr::all_of(unname(p4_vars)), ~ sum(.x)))

# percentages of the total, with a coalition column (everyone except non-Hispanic White alone)
comp <- bind_rows(sw, dplyr::select(dist_comp, -DISTRICT))
district_labels <- c("Statewide", as.character(dist_comp$DISTRICT))
pct <- comp |>
  mutate(Coalition = round((pop_vap - P4_005N) / pop_vap * 100, 1),
         dplyr::across(dplyr::all_of(unname(p4_vars)), ~ round(.x / pop_vap * 100, 1)))

desc_tab <- cbind(District = district_labels,
                  dplyr::select(pct, dplyr::all_of(unname(p4_vars)), Coalition))
names(desc_tab) <- c("District", names(p4_vars), "Coalition")
save_output_table(desc_tab, "Race and ethnicity of the voting-age population by district")

# highlight the column for the analysis group
group_col <- if (tolower(params$group) == "hispanic") "Hispanic" else
             if (tolower(params$group) == "coalition") "Coalition" else
             paste0(params$group, " (NH)")
hl <- match(group_col, names(desc_tab))

kb <- desc_tab |>
  kbl(caption = paste0("Race and ethnicity of the voting-age population by district (%), ",
                       params$state, ". NH is non-Hispanic alone; Coalition is all but White (NH).")) |>
  style_tbl() |>
  row_spec(1, bold = TRUE, background = "#EAF0F1")
if (!is.na(hl)) kb <- kb |> column_spec(hl, bold = TRUE, background = "#FDEBD0")
kb |> scroll_box(width = "100%")
```

Table \@ref(tab:desc-comp-total) repeats the district composition on the total population of all
ages rather than the voting-age population, using table P2. A group's share can shift between the
two bases, so both are reported. Read it the same way, comparing each district row to the Statewide
row on top.

```{r desc-comp-total, echo = FALSE, warning = FALSE, message = FALSE}
# same categories as the voting-age table, but on the total population (all ages) from
# table P2, disaggregated to districts by the same area-weighted routine
p2_vars <- c(`White (NH)` = "P2_005N", `Black (NH)` = "P2_006N",
             `AIAN (NH)` = "P2_007N", `Asian (NH)` = "P2_008N",
             `NHPI (NH)` = "P2_009N", `Other (NH)` = "P2_010N",
             `Two+ (NH)` = "P2_011N", `Hispanic` = "P2_002N")

bg_comp_tot <- blockgroups |>
  dplyr::filter(!st_is_empty(geom)) |>
  dplyr::select(pop_total = P2_001N, dplyr::all_of(unname(p2_vars)))

dist_comp_tot <- disaggregate_to_fragments(bg_comp_tot, districts,
                                           c("pop_total", unname(p2_vars))) |>
  aggregate_to_target("DISTRICT", c("pop_total", unname(p2_vars))) |>
  arrange(DISTRICT)

# statewide totals on the total population
sw_tot <- blockgroups |> st_drop_geometry() |>
  dplyr::summarize(pop_total = sum(P2_001N),
                   dplyr::across(dplyr::all_of(unname(p2_vars)), ~ sum(.x)))

comp_tot <- bind_rows(sw_tot, dplyr::select(dist_comp_tot, -DISTRICT))
district_labels_tot <- c("Statewide", as.character(dist_comp_tot$DISTRICT))
pct_tot <- comp_tot |>
  mutate(Coalition = round((pop_total - P2_005N) / pop_total * 100, 1),
         dplyr::across(dplyr::all_of(unname(p2_vars)), ~ round(.x / pop_total * 100, 1)))

desc_tab_tot <- cbind(District = district_labels_tot,
                      dplyr::select(pct_tot, dplyr::all_of(unname(p2_vars)), Coalition))
names(desc_tab_tot) <- c("District", names(p2_vars), "Coalition")
save_output_table(desc_tab_tot, "Race and ethnicity of the total population by district")

# highlight the same analysis-group column as the voting-age table (group_col from above)
hl_t <- match(group_col, names(desc_tab_tot))
kb_t <- desc_tab_tot |>
  kbl(caption = paste0("Race and ethnicity of the total population (all ages) by district (%), ",
                       params$state, ". NH is non-Hispanic alone; Coalition is all but White (NH).")) |>
  style_tbl() |>
  row_spec(1, bold = TRUE, background = "#EAF0F1")
if (!is.na(hl_t)) kb_t <- kb_t |> column_spec(hl_t, bold = TRUE, background = "#FDEBD0")
kb_t |> scroll_box(width = "100%")
```

Table \@ref(tab:desc-statewide) shows the statewide size of each group on two population bases, the
voting-age population used in the analysis and the total population of all ages. A group's share can
differ between the two, so both are reported.

```{r desc-statewide, echo = FALSE, warning = FALSE, message = FALSE}
p4s <- get_decennial(geography = "state", state = params$state, sumfile = "pl",
                     table = "P4", year = 2020, output = "wide")
p2s <- get_decennial(geography = "state", state = params$state, sumfile = "pl",
                     table = "P2", year = 2020, output = "wide")

desc_statewide <- data.frame(
  Group = c("White (NH)", "Black (NH)", "AIAN (NH)", "Asian (NH)", "NHPI (NH)",
            "Other (NH)", "Two+ (NH)", "Hispanic", "Coalition (all but White NH)", "Total"),
  vap = c(p4s$P4_005N, p4s$P4_006N, p4s$P4_007N, p4s$P4_008N, p4s$P4_009N,
          p4s$P4_010N, p4s$P4_011N, p4s$P4_002N, p4s$P4_001N - p4s$P4_005N, p4s$P4_001N),
  tot = c(p2s$P2_005N, p2s$P2_006N, p2s$P2_007N, p2s$P2_008N, p2s$P2_009N,
          p2s$P2_010N, p2s$P2_011N, p2s$P2_002N, p2s$P2_001N - p2s$P2_005N, p2s$P2_001N),
  stringsAsFactors = FALSE) |>
  mutate(`VAP` = format(round(vap), big.mark = ","),
         `VAP %` = round(vap / p4s$P4_001N * 100, 1),
         `Total pop` = format(round(tot), big.mark = ","),
         `Total pop %` = round(tot / p2s$P2_001N * 100, 1)) |>
  dplyr::select(Group, `VAP`, `VAP %`, `Total pop`, `Total pop %`)

save_output_table(desc_statewide, "Statewide composition by group on two population bases")

# highlight the analysis group's row (group_col comes from the desc-comp chunk above;
# the coalition row carries a longer label, so match it by prefix)
grp_row_sw <- which(desc_statewide$Group == group_col |
                    (group_col == "Coalition" & startsWith(desc_statewide$Group, "Coalition")))
kb_sw <- desc_statewide |>
  kbl(caption = paste0("Statewide composition by group on two population bases, ", params$state,
                       " (VAP is voting age, Total pop is all ages).")) |>
  style_tbl() |>
  row_spec(nrow(desc_statewide), bold = TRUE)
if (length(grp_row_sw)) kb_sw <- kb_sw |> row_spec(grp_row_sw, bold = TRUE, background = "#FDEBD0")
kb_sw
```

### 3.3.4 Voting summary by district

`r if (have_votes) "Table \\@ref(tab:vote-summary) summarizes the two-party vote by district, the raw counts that parallel the racial composition above. The focal party is the statewide underdog by default, or the party forced in the header." else "Precinct vote data was not loaded, so there is no voting summary for this run."`

```{r vote-summary, echo = FALSE, eval = have_votes}
# two-party votes by district, from the same area-weighting used elsewhere (block groups to
# districts), so these totals match the ones the analysis uses. A statewide row sits on top,
# parallel to the race table, with the other party's votes shown alongside the focal party's.
vote_by_dist <- disaggregate_to_fragments(blockgroups_calc, districts, c("totV", "mpV")) |>
  aggregate_to_target("DISTRICT", c("totV", "mpV")) |>
  arrange(DISTRICT) |>
  dplyr::transmute(District = as.character(DISTRICT), totV, mpV)
vote_sw <- dplyr::summarize(vote_by_dist, District = "Statewide",
                            totV = sum(totV), mpV = sum(mpV))
vote_tab <- bind_rows(vote_sw, vote_by_dist) |>
  dplyr::transmute(District,
                   `Total votes` = round(totV),
                   `Focal party votes` = round(mpV),
                   `Other party votes` = round(totV - mpV),
                   `Focal party %` = round(mpV / totV * 100, 1))
save_output_table(vote_tab, "Two-party voting summary by district")
vote_tab |>
  kbl(caption = paste0("Two-party voting summary by district, ", params$state,
                       ". The first row is the statewide total.")) |>
  style_tbl() |>
  row_spec(1, bold = TRUE, background = "#EAF0F1")
```

### 3.3.5 Dot density map (where the group lives)

A dot-density map (Figure \@ref(fig:map-dots-group)) shows where people live, which the tables above cannot show. Each dot is a
fixed number of people, the group set by `params$group` in one color and everyone else in grey.
This is a descriptive picture, not a result, but it matters because packing and cracking depend on
where people already live. To read it, look for tight clusters of the group's color, which mark
where it concentrates, often a city or a historic community, and compare those clusters to the
district lines. When a district line wraps tightly around a cluster, that is a visual sign of
packing, and when a line cuts a cluster into several districts, that is a visual sign of cracking.


This map is shown only at coarser resolutions. At block level there are too many tiny
units to sample and render, so it is skipped automatically.

```{r map-dots-setup, include = FALSE}
# dot density is only feasible at coarse units; skip it automatically for block-level data
do_dots <- geo != "block"
do_vote_dots <- do_dots && have_votes   # voting dot map also needs precinct data
set.seed(1)
dot_people <- 1000   # people per dot (voting age); raise this to speed up rendering
bg_dots <- blockgroups_calc |> dplyr::filter(!st_is_empty(geom))

# helper: build one point layer per category column, sampled in proportion to its count
sample_dots <- function(bg, cols) {
  parts <- lapply(names(cols), function(nm) {
    n   <- round(pmax(st_drop_geometry(bg)[[cols[[nm]]]], 0) / dot_people)
    sfc <- st_sample(bg, size = n)
    if (length(sfc) == 0) NULL else st_sf(kind = nm, geometry = sfc)
  })
  out <- do.call(rbind, parts[!vapply(parts, is.null, logical(1))])
  out$kind <- factor(out$kind, levels = names(cols))
  out
}

# public folder for the shareable, census-derived dot layer (per state). The census inputs
# are public, so this derived layer can live outside the private folder; the vote dot layer
# stays in the private result folder because it is built from non-redistributable RDH data.
output_public <- here("data", "derived", "public", state_slug)
dir.create(output_public, recursive = TRUE, showWarnings = FALSE)
```

```{r map-dots-group, eval = do_dots, echo = FALSE, fig.width = map_w, fig.height = map_w * map_asp + 0.6, fig.cap = paste0("Dot density of ", params$group, " and everyone else, ", params$state, ".")}
# reuse the minority and pop_basis already computed on blockgroups_calc
bg_dots$grp <- bg_dots$minority
bg_dots$oth <- pmax(bg_dots$pop_basis - bg_dots$minority, 0)

dot_layer <- sample_dots(bg_dots, setNames(c("grp", "oth"), c(params$group, "Other")))
group_slug <- gsub("[^A-Za-z0-9]+", "_", params$group)
st_write(dot_layer, file.path(output_public, paste0("dot_density_", group_slug, ".gpkg")),
         delete_dsn = TRUE, quiet = TRUE)

ggplot() +
  geom_sf(data = dot_layer, aes(color = kind, size = kind), alpha = 0.35) +
  geom_sf(data = districts, fill = NA, color = "grey20", linewidth = 0.45) +
  scale_color_manual(values = setNames(c("#B34700", "grey75"), c(params$group, "Other")),
                     name = NULL) +
  scale_size_manual(values = setNames(c(0.7, 0.3), c(params$group, "Other")), guide = "none") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(subtitle = paste0("1 dot = ", dot_people, " people (voting age)")) +
  theme_void(base_size = 14) +
  theme(legend.position = "bottom", plot.margin = margin(0, 0, 0, 0))
```

When precinct data is present, the same dot-density view for the focal party vote shows where each
party's votes concentrate, alongside the group's population above.

```{r map-dots-vote, eval = do_vote_dots, echo = FALSE, fig.width = map_w, fig.height = map_w * map_asp + 0.6, fig.cap = paste0("Dot density of the ", fp_disp, " vote and the other party, ", params$state, ".")}
bg_dots$mp_vote  <- bg_dots$mpV
bg_dots$oth_vote <- pmax(bg_dots$totV - bg_dots$mpV, 0)

vote_dots <- sample_dots(bg_dots, setNames(c("mp_vote", "oth_vote"),
                                           c(fp_disp, "Other party")))
st_write(vote_dots, file.path(output_dir, "dot_density_vote.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)

ggplot() +
  geom_sf(data = vote_dots, aes(color = kind, size = kind), alpha = 0.35) +
  geom_sf(data = districts, fill = NA, color = "grey20", linewidth = 0.45) +
  scale_color_manual(values = setNames(c("#CA0020", "grey75"),
                                       c(fp_disp, "Other party")), name = NULL) +
  scale_size_manual(values = setNames(c(0.7, 0.3), c(fp_disp, "Other party")), guide = "none") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(subtitle = paste0("1 dot = ", dot_people, " votes")) +
  theme_void(base_size = 14) +
  theme(legend.position = "bottom", plot.margin = margin(0, 0, 0, 0))
```

# 4 Analysis

## 4.1 District composition by area-weighted reaggregation

We estimate the minority share of each district by area-weighted reaggregation of the
block groups. Intersecting the block groups with the districts splits each block group
into fragments, one per district it touches. We allocate each block group's population
to its fragments in proportion to area, then sum the fragments by district. We keep the
fragments (`race_fragments`) because later steps reuse them. The intersection is made
robust by keeping only polygons, repairing invalid geometries, dropping empty results,
and casting to single-part polygons. These two steps are wrapped in the reusable
`disaggregate_to_fragments` and `aggregate_to_target` functions, defined earlier so that
the descriptive composition table uses the same routine.

```{r district-race, warning = FALSE}
# disaggregate block groups into district fragments, then aggregate back to districts.
# race_fragments keeps the fragment geometry because the compact-population-center step reuses it.
race_fragments <- disaggregate_to_fragments(blockgroups_calc, districts,
                    c("pop_basis", "minority", "minority_tot", "pop_total", "totV", "mpV"))

district_estimates <- race_fragments |>
  aggregate_to_target("DISTRICT",
                      c("pop_basis", "minority", "minority_tot", "pop_total", "totV", "mpV")) |>
  rename(bgTotal = pop_basis, bgMinority = minority,
         bgMinorityTot = minority_tot, bgTotalPop = pop_total)

districts_join <- districts |>
  left_join(district_estimates, by = "DISTRICT") |>
  mutate(pctMinoritybg = round(bgMinority / bgTotal * 100, 1),
         pctMinoritybg_tot = round(bgMinorityTot / bgTotalPop * 100, 1),
         pctMP = round(mpV / totV * 100, 1))
```

## 4.2 Convex hull composition

The convex hull of a district is the smallest convex shape that contains it, like a
rubber band stretched around the district. We build one hull per district and estimate
the minority share inside it by the same area-weighted reaggregation, this time
intersecting the block groups with the hulls. When precinct data is present we do the
same for votes.

```{r convex-hull, warning = FALSE}
d_ch <- districts |>
  dplyr::select(DISTRICT) |>
  st_convex_hull() |>
  mutate(chArea = st_area(geom))

d_ch_est <- disaggregate_to_fragments(blockgroups_calc, d_ch, c("pop_basis", "minority")) |>
  aggregate_to_target(c("DISTRICT", "chArea"), c("pop_basis", "minority")) |>
  rename(chTotal = pop_basis, chMinority = minority)

# save the constructed hull shapes with their aggregated data
d_ch |> left_join(d_ch_est, by = c("DISTRICT", "chArea")) |>
  st_write(file.path(output_dir, "hull_shapes.gpkg"), delete_dsn = TRUE, quiet = TRUE)

if (have_votes) {
  d_ch_estV <- disaggregate_to_fragments(precincts, d_ch, c("totV", "mpV")) |>
    aggregate_to_target("DISTRICT", c("totV", "mpV")) |>
    rename(chtotV = totV, chmpV = mpV)
}
```

```{r map-hull, echo = FALSE, fig.width = map_w, fig.height = map_w * bbox_asp(d_ch), fig.cap = paste0("Each district in ", params$state, " with its convex hull.")}
ggplot() +
  geom_sf(data = districts, fill = "#C1E3CD", color = "#006d2c", alpha = 0.8, linewidth = 0.2) +
  geom_sf(data = d_ch, fill = "#F0FFFF", color = "#6EA6C3", alpha = 0.4, linewidth = 0.5) +
  geom_sf_text(data = districts, aes(label = DISTRICT), size = 4, fontface = "bold", color = "grey15") +
  theme_void()
```

## 4.3 Minimum bounding circle composition

The minimum bounding circle is the smallest circle that contains a district. Because it
is a shape operation, we compute it in a projected coordinate system rather than in
geographic coordinates. We pick the NAD83 UTM zone automatically from the data, build
the circle, and transform back to EPSG:4269. We then estimate the minority and vote
share inside each circle in the same way as for the hulls.

```{r mbc, warning = FALSE}
# auto-select the NAD83 UTM zone from the districts' centroid longitude
ctr <- suppressWarnings(st_coordinates(st_centroid(st_union(districts))))
mbc_crs <- 26900L + as.integer(floor((ctr[1] + 180) / 6) + 1)

d_mbc <- districts |>
  dplyr::select(DISTRICT) |>
  st_transform(3395) |>
  lwgeom::st_minimum_bounding_circle() |>
  st_transform(4269) |>
  mutate(mbcArea = st_area(geom))

d_mbc_est <- disaggregate_to_fragments(blockgroups_calc, d_mbc, c("pop_basis", "minority")) |>
  aggregate_to_target(c("DISTRICT", "mbcArea"), c("pop_basis", "minority")) |>
  rename(mbcTotal = pop_basis, mbcMinority = minority)

# save the constructed minimum bounding circle shapes with their aggregated data
d_mbc |> left_join(d_mbc_est, by = c("DISTRICT", "mbcArea")) |>
  st_write(file.path(output_dir, "mbc_shapes.gpkg"), delete_dsn = TRUE, quiet = TRUE)

if (have_votes) {
  d_mbc_estV <- disaggregate_to_fragments(precincts, d_mbc, c("totV", "mpV")) |>
    aggregate_to_target("DISTRICT", c("totV", "mpV")) |>
    rename(mbctotV = totV, mbcmpV = mpV)
}
```

```{r map-mbc, echo = FALSE, fig.width = map_w, fig.height = map_w * bbox_asp(d_mbc), fig.cap = paste0("Each district in ", params$state, " with its minimum bounding circle.")}
ggplot() +
  geom_sf(data = districts, fill = "#C1E3CD", color = "#006d2c", alpha = 0.8, linewidth = 0.2) +
  geom_sf(data = d_mbc, fill = "#F0FFFF", color = "#6EA6C3", alpha = 0.4, linewidth = 0.5) +
  geom_sf_text(data = districts, aes(label = DISTRICT), size = 4, fontface = "bold", color = "grey15") +
  theme_void()
```

## 4.4 Compactness scores and the ΔC computation

We now compare each district to its compact reference shapes. Three compactness scores
describe how regular a district's shape is. The Polsby-Popper score uses the district
area and perimeter, the convex hull ratio compares the district area to its hull area,
and the Reock score compares the district area to its minimum bounding circle. The
ΔC scores compare composition instead of shape. `chDiff` is the district's minority
percentage minus the minority percentage of its convex hull, and `mbcDiff` is the same
for the minimum bounding circle. A positive ΔC means concentration and a negative ΔC
means dilution. When precinct data is present we compute the same differences for
the focal party vote as `chDiffV` and `mbcDiffV`.

```{r cdiff}
districts_results <- districts_join |>
  left_join(d_ch_est, by = "DISTRICT") |>
  mutate(chPct = round(chMinority / chTotal * 100, 1),
         chDiff = pctMinoritybg - chPct, chAbsDiff = abs(chDiff)) |>
  left_join(d_mbc_est, by = "DISTRICT") |>
  mutate(mbcPct = round(mbcMinority / mbcTotal * 100, 1),
         mbcDiff = pctMinoritybg - mbcPct, mbcAbsDiff = abs(mbcDiff))

if (have_votes) {
  districts_results <- districts_results |>
    left_join(d_ch_estV, by = "DISTRICT") |>
    mutate(chPctV = round(chmpV / chtotV * 100, 1),
           chDiffV = pctMP - chPctV, chAbsDiffV = abs(chDiffV)) |>
    left_join(d_mbc_estV, by = "DISTRICT") |>
    mutate(mbcPctV = round(mbcmpV / mbctotV * 100, 1),
           mbcDiffV = pctMP - mbcPctV, mbcAbsDiffV = abs(mbcDiffV))
}

# Polsby-Popper, hull ratio, and Reock, all from geodesic area and perimeter on EPSG 4269
# (with s2 off, sf uses lwgeom's geodesic measures, so no projected CRS is needed).
districts_results <- districts_results |>
  mutate(darea  = st_area(geom),
         dperim = st_length(st_cast(geom, "MULTILINESTRING")),
         compact_shp  = round(as.numeric((4 * pi * darea) / dperim^2), 2),
         compact_hull = round(as.numeric(darea / chArea), 2),
         compact_circ = round(as.numeric(darea / mbcArea), 2))
```

Table \@ref(tab:compactness-table) reports the three compactness scores for each district as a
shape reference. They describe only how regular each district's outline is, which is a separate
question from the composition that the ΔC scores address. A score near 1 is compact and a
score near 0 is stringy.

```{r compactness-table, echo = FALSE}
compact_tab <- districts_results |> st_drop_geometry() |>
  dplyr::select(District = DISTRICT,
                `Polsby-Popper` = compact_shp,
                `Convex hull ratio` = compact_hull,
                `Reock` = compact_circ) |>
  arrange(District)
save_output_table(compact_tab, "Shape-compactness scores by district")
compact_tab |>
  kbl(caption = paste0("Shape-compactness scores by district, ", params$state,
                       " (near 1 is compact, near 0 is stringy).")) |>
  style_tbl()
```

## 4.5 Compact population center composition

We add a third compact reference shape. For each district we find the
population-weighted mean center of its residents and grow a compact district outward
from that center, adding the nearest population until it reaches the target size. The
target size is the state's total population divided by the number of districts, because
redistricting equalizes total population. Centers and distances are computed in a
projected coordinate system so that distances are in meters. As with the other shapes,
composition is measured with the `r basis_word`.

```{r save-load-race-fragments, warning = FALSE}
race_fragments |> st_write(
  file.path(output_dir, "race_frags.gpkg"),
  append = FALSE
)

```


```{r cpc-build, warning = FALSE}
n_districts <- dplyr::n_distinct(districts$DISTRICT)
idealpop <- sum(blockgroups_calc$pop_total) / n_districts

# fragment centroids: projected (meters) for the population-weighted mean center, and
# geographic (lon/lat) for the geodesic distance used to grow the compact district
frag_cent <- race_fragments |> st_centroid()
frag_ll   <- st_coordinates(frag_cent)                    # lon, lat (EPSG 4269)
frag_xy   <- st_coordinates(st_transform(frag_cent, mbc_crs))
frag_tab  <- st_drop_geometry(frag_cent) |>
  mutate(x = frag_xy[, 1], y = frag_xy[, 2],
         lon = frag_ll[, 1], lat = frag_ll[, 2])

# population-weighted mean center of each district (projected meters), then its lon/lat
centers <- frag_tab |>
  group_by(DISTRICT) |>
  summarize(cx = sum(x * pop_total) / sum(pop_total),
            cy = sum(y * pop_total) / sum(pop_total), .groups = "drop") |>
  rename(cpcDistrict = DISTRICT)
centers_ll <- st_as_sf(centers, coords = c("cx", "cy"), crs = mbc_crs) |>
  st_transform(4269) |> st_coordinates()
centers <- centers |> mutate(clon = centers_ll[, 1], clat = centers_ll[, 2])

# for each center, take the nearest fragments until the target population is reached.
# distance is the geodesic (ellipsoidal) length between the fragment and the district
# center, so it is not distorted by the map projection
cpc_assign <- frag_tab |>
  dplyr::select(frag_id, DISTRICT, lon, lat, pop_total, pop_basis, minority, minority_tot, totV, mpV) |>
  cross_join(centers) |>
  mutate(dist = geosphere::distGeo(cbind(lon, lat), cbind(clon, clat))) |>
  arrange(cpcDistrict, dist) |>
  group_by(cpcDistrict) |>
  mutate(cumpop = cumsum(pop_total)) |>
  dplyr::filter(cumpop <= idealpop) |>
  ungroup()
```

Create one spatial layer of locations and their status *vis-a-vis* cpc's.

```{r cpc-fragments, warning = FALSE}
# join cpc assignment table to race_fragments spatial features
cpc_fragments <- race_fragments |>
  left_join(dplyr::select(cpc_assign, frag_id, cpcDistrict), by = "frag_id")

# identity fragments are inside of both the district and it's cpc
cpc_identity <- cpc_fragments |> filter(DISTRICT == cpcDistrict) |> 
  mutate(cpc_type = 'identity')

# captured fragments are inside the district, but not it's cpc, as if the
# compact district boundaries were stretched out to capture these areas
cpc_capture <- race_fragments |> 
  anti_join(st_drop_geometry(cpc_identity), by = "frag_id") |> 
  mutate(cpc_type = 'captured')

# excluded fragments are outside the district, but inside it's cpc, as if
# the compact district boundaries were shrunk / contorted to exclude these areas
cpc_excluded <- cpc_fragments |> 
  filter(DISTRICT != cpcDistrict) |> 
  mutate(cpc_type = 'excluded') 

# count how many cpc's overlapped
cpc_excluded_n <- cpc_excluded |>
  st_drop_geometry() |> 
  group_by(frag_id) |> 
  summarize(excluded_count = n(), .groups ="drop")

# join the count back to the cpc_excluded fragments
cpc_excluded <- cpc_excluded |> 
  left_join(cpc_excluded_n, by = "frag_id")

# merge all three cases of CPCs together and save
cpc_fragments <- bind_rows(cpc_identity, cpc_capture, cpc_excluded)

cpc_fragments |> st_write(file.path(output_dir, "cpc_fragments.gpkg"),
                          append = FALSE)
```

Summarize CPC results at the district level

```{r cpc-summary, warning = FALSE}
# composition inside each compact population-center district (voting-age)
cpc_est <- cpc_assign |>
  group_by(cpcDistrict) |>
  summarize(cpcTotal = sum(pop_basis),
            cpcMinority = sum(minority),
            cpcMinorityTot = sum(minority_tot),
            cpcIdealPop = sum(pop_total),
            cpctotV = sum(totV),
            cpcmpV = sum(mpV)) |>
  mutate(cpcPct = round(cpcMinority / cpcTotal * 100, 1),
         cpcPctTot = round(cpcMinorityTot / cpcIdealPop * 100, 1),
         cpcPctV = round(cpcmpV / cpctotV * 100, 1))

cpc_cases <- cpc_fragments |>
  group_by(DISTRICT, cpcDistrict, cpc_type) |>
  summarize(cpcTotal = sum(pop_basis),
            cpcTargetPop = sum(pop_total),
            cpcMinority = sum(minority),
            cpcMinorityTot = sum(minority_tot),
            cpctotV = sum(totV),
            cpcmpV = sum(mpV)) |>
  ungroup()

cpc_cases |> st_write(file.path(output_dir, "cpc_cases.gpkg"),
                      append = FALSE)

cpcd_identity <- cpc_cases |> 
  filter(cpc_type == 'identity') |> 
  st_drop_geometry() |> 
  mutate(cpcidTotal = cpcTotal,
         cpcidMinority = cpcMinority,
         cpcidMinorityTot = cpcMinorityTot,
         cpcidIdealPop = cpcTargetPop,
         cpcidPctMinority = round(cpcidMinority / cpcidTotal * 100, 1),
         cpcidPctMinorityTot = round(cpcidMinorityTot / cpcidIdealPop * 100, 1),
         cpcidtotV = cpctotV,
         cpcidmpV = cpcmpV,
         cpcidPctV = round(cpcidmpV / cpcidtotV * 100, 1)) |>
  select(DISTRICT, starts_with("cpcid"))

cpcd_captured <- cpc_cases |> 
  filter(cpc_type == 'captured') |> 
  st_drop_geometry() |> 
  mutate(cpccaTotal = cpcTotal,
         cpccaMinority = cpcMinority,
         cpccaMinorityTot = cpcMinorityTot,
         cpccaIdealPop = cpcTargetPop,
         cpccaPctMinority = round(cpccaMinority / cpccaTotal * 100, 1),
         cpccaPctMinorityTot = round(cpccaMinorityTot / cpccaIdealPop * 100, 1),
         cpccatotV = cpctotV,
         cpccampV = cpcmpV,
         cpccaPctV = round(cpccampV / cpccatotV * 100, 1)) |>
  select(DISTRICT, starts_with("cpcca"))

cpcd_excluded <- cpc_cases |> 
  filter(cpc_type == 'excluded') |> 
  st_drop_geometry() |> 
  group_by(cpcDistrict) |> 
  summarize(cpcexTotal = sum(cpcTotal),
            cpcexMinority = sum(cpcMinority),
            cpcexMinorityTot = sum(cpcMinorityTot),
            cpcexIdealPop = sum(cpcTargetPop),
            cpcextotV = sum(cpctotV),
            cpcexmpV = sum(cpcmpV)) |>
  mutate(DISTRICT = cpcDistrict,
         cpcexPctMinority = round(cpcexMinority / cpcexTotal * 100, 1),
         cpcexPctMinorityTot = round(cpcexMinorityTot / cpcexIdealPop * 100, 1),
         cpcexPctV = round(cpcexmpV / cpcextotV * 100, 1)) |>
  select(DISTRICT, starts_with("cpcex"))

# dissolve the assigned fragments into compact population-center district shapes
cpc_districts <- bind_rows(cpc_excluded, cpc_identity) |>
  group_by(cpcDistrict) |>
  summarize() |> 
  left_join(cpc_est, by="cpcDistrict")

cpc_districts |> st_write(file.path(output_dir, "cpc_shapes.gpkg"),
                      append = FALSE)

districts_results <- districts_results |>
  left_join(cpc_est, by = c("DISTRICT" = "cpcDistrict")) |>
  left_join(cpcd_identity, by = "DISTRICT") |> 
  left_join(cpcd_captured, by = "DISTRICT") |> 
  left_join(cpcd_excluded, by = "DISTRICT")

districts_results <- districts_results |>
  mutate(cpcDiff = pctMinoritybg - cpcPct, cpcAbsDiff = abs(cpcDiff),
         cpcIdentityPct = cpcidTotal / bgTotal * 100,
         cpcCapturePctDiff = cpccaPctMinority - cpcidPctMinority,
         cpcExcludePctDiff = cpcexPctMinority - cpcidPctMinority,
         # total-population versions, used by the CPC decomposition figure (minority basis)
         cpcDiffTot = pctMinoritybg_tot - cpcPctTot,
         cpcIdentityPctTot = cpcidIdealPop / bgTotalPop * 100,
         cpcDiffV = pctMP - cpcPctV,
         cpcAbsDiffV = abs(cpcDiffV),
         cpcIdentityPctV = cpcidtotV / totV * 100,
         cpcCapturePctDiffV = cpccaPctV - cpcidPctV,
         cpcExcludePctDiffV = cpcexPctV - cpcidPctV,
         )
```


With the compact population center built for every district, we read its results in three linked
views, first for the group population and then, when precinct data is present, for the focal party
vote. The three views are a scatter of ΔC against how compact each district already is, a dumbbell
that splits each district into the population it reaches out to capture and the population it leaves
out, and a map that locates those results in space.

Figure \@ref(fig:scatterplot-cpcs) places each district by the share of its population inside the
overlap region, the part that falls in both the actual district and its equal-population circle, on
the horizontal axis, against its ΔC on the vertical axis. A district high on the vertical axis holds
a larger group share than its compact center, which reads as concentration, and one low on it reads
as dilution. The horizontal axis says how much of the district is already compact, so a point far to
the right is mostly its own overlap region while a point to the left has been reshaped more.

```{r scatterplot-cpcs, echo = FALSE, fig.width = 7, fig.height = 6, fig.cap = paste0("CPC overlap share versus ΔC by district, ", params$state, ".")}
ggplot(districts_results, aes(x = cpcIdentityPct, y = cpcDiff)) +
  geom_hline(yintercept = 0, color = "grey70") +
  geom_point(aes(fill = ifelse(cpcDiff >= 0, "Concentration", "Dilution")),
             shape = 21, size = 9, color = "grey30", stroke = 0.6) +
  geom_text(aes(label = DISTRICT), color = "white", fontface = "bold", size = 4) +
  scale_fill_manual(values = c("Concentration" = "#DC6D85", "Dilution" = "#6EA6C3"), name = NULL) +
  labs(x = "CPC overlap share of VAP (%)", y = "ΔC (percentage points)") +
  theme_bw(base_size = 14) +
  theme(legend.position = "top")
```

```{r cpc-dumbbell-setup, include = FALSE}
# Reusable dumbbell for the CPC decomposition, used by the race and the vote figures.
# Each district is one row. The overlap region (inside both the district and its equal-population
# circle) is fixed at 0 on a common vertical line, and the two markers are each region's share
# relative to that overlap: a circle for the captured region (population the
# district reaches out to pull in) and a diamond for the excluded region (population the compact
# circle would have held but the district leaves out). The whole row (line, both markers, and
# the district label) is colored by direction: red for concentration (positive ΔC) and blue for
# dilution (negative ΔC). Marker area is the region's size (population or votes), and within a
# row the circle and diamond positions show how the ΔC came about. Districts are ordered by ΔC,
# and each region's actual share is printed in black just outside its marker.
make_cpc_dumbbell <- function(overlap_col, captured_col, excluded_col,
                              cap_size_col, exc_size_col, diff_col, value_label) {
  df <- districts_results |> st_drop_geometry() |>
    dplyr::transmute(DISTRICT,
                     overlap  = .data[[overlap_col]],
                     captured = .data[[captured_col]],
                     excluded = .data[[excluded_col]],
                     cap_size = .data[[cap_size_col]],
                     exc_size = .data[[exc_size_col]],
                     dC       = .data[[diff_col]]) |>
    arrange(dC) |>
    mutate(rank = row_number(),
           direction = ifelse(dC >= 0, "Concentration", "Dilution"),
           dir_col = ifelse(dC >= 0, "#C0405E", "#3E7EA0"),
           row_label = paste0("<span style='color:", dir_col, "'>District ", DISTRICT,
                              " (", sprintf("%+.1f", dC), ")</span>"),
           cap_dev = captured - overlap,   # each region's share relative to the overlap region
           exc_dev = excluded - overlap,
           cap_right = cap_dev >= exc_dev,
           cap_x = cap_dev + ifelse(cap_right,  4, -4),   # print the actual share just outside
           cap_h = ifelse(cap_right, 0, 1),               # the marker, on the outboard side
           exc_x = exc_dev + ifelse(cap_right, -4,  4),
           exc_h = ifelse(cap_right, 1, 0))
  ggplot(df, aes(y = rank)) +
    geom_vline(xintercept = 0, color = "grey45", linewidth = 0.5) +
    geom_segment(aes(x = exc_dev, xend = cap_dev, y = rank, yend = rank, color = direction),
                 linewidth = 1) +
    geom_point(aes(x = exc_dev, size = exc_size, color = direction,
                   shape = "Excluded (left out)")) +
    geom_point(aes(x = cap_dev, size = cap_size, color = direction,
                   shape = "Captured (pulled in)")) +
    geom_text(aes(x = cap_x, label = round(captured), hjust = cap_h),
              size = 3.8, color = "black") +
    geom_text(aes(x = exc_x, label = round(excluded), hjust = exc_h),
              size = 3.8, color = "black") +
    scale_color_manual(values = c("Concentration" = "#C0405E", "Dilution" = "#3E7EA0"),
                       name = NULL) +
    scale_shape_manual(values = c("Captured (pulled in)" = 16, "Excluded (left out)" = 18),
                       breaks = c("Captured (pulled in)", "Excluded (left out)"), name = NULL) +
    scale_size_area(max_size = 12, guide = "none") +
    scale_y_reverse(breaks = df$rank, labels = df$row_label) +
    labs(x = paste0(value_label, " minus the overlap share (percentage points, 0 = overlap)"),
         y = "District (ΔC), ordered by ΔC") +
    theme_bw(base_size = 15) +
    theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
          axis.text.y = ggtext::element_markdown(),
          legend.position = "bottom", legend.box = "vertical", legend.spacing.y = unit(2, "pt")) +
    guides(color = guide_legend(order = 1, override.aes = list(shape = NA, linewidth = 1.4)),
           shape = guide_legend(order = 2, override.aes = list(size = 4, color = "grey30")))
}
```

Figure \@ref(fig:cpc-viz) is that dumbbell for the group population, measured on the total population of
all ages. Each district is one row, and the grey vertical line at 0 is the overlap region, the part
inside both the actual district and its equal-population circle, the same reference for every
district so the rows line up. The horizontal axis is each region's share minus the overlap share, so 0
means the same as the overlap, a marker to the right is more group-heavy than the overlap, and one to
the left is less. The captured region is a circle and the excluded region is a diamond. The whole row, the line, both markers, and the district
label, is red when the district concentrates the group, a positive ΔC, and blue when it dilutes it, a
negative ΔC. Within a row the circle and diamond positions show how that came about, whether by reaching
out to capture group-heavy population or by carving out group-heavy population. Marker area is the
region's total population, and each region's actual share is printed in black just outside its marker.

```{r cpc-viz, echo = FALSE, warning = FALSE, fig.height = max(5, 0.9 * n_districts), fig.width = 10, fig.cap = paste0("CPC decomposition of each district into overlap, captured, and excluded regions by the group's share of total population, ", params$state, ". Read together with the ΔC map that follows.")}
make_cpc_dumbbell("cpcidPctMinorityTot", "cpccaPctMinorityTot", "cpcexPctMinorityTot",
                  "cpccaIdealPop", "cpcexIdealPop", "cpcDiffTot",
                  "Minority share")
```

Figure \@ref(fig:cpc-viz-map) is the locator for those bars. It shades each district on the state map
by its total-population ΔC, red for concentration and blue for dilution, so the abstract bars above
have a geographic place and you can see where the concentrated and diluted districts sit.

```{r cpc-viz-map, echo = FALSE, warning = FALSE, fig.width = map_w, fig.height = map_w * map_asp + 1.0, fig.cap = paste0("Compact population-center ΔC by district, ", params$state, ". The locator for the CPC decomposition above.")}
cpc_map_df <- districts |> dplyr::select(DISTRICT) |>
  left_join(dplyr::select(st_drop_geometry(districts_results), DISTRICT, cpcDiffTot),
            by = "DISTRICT")
cpc_map_lim <- max(abs(cpc_map_df$cpcDiffTot), na.rm = TRUE)
ggplot(cpc_map_df) +
  geom_sf(aes(fill = cpcDiffTot), color = "grey40", linewidth = 0.2) +
  geom_sf_text(aes(label = DISTRICT), size = 4.5, fontface = "bold", color = "grey15") +
  scale_fill_gradient2(low = "#6EA6C3", mid = "white", high = "#DC6D85", midpoint = 0,
                       limits = c(-cpc_map_lim, cpc_map_lim),
                       breaks = c(-cpc_map_lim, 0, cpc_map_lim),
                       labels = c("Dilution", "0", "Concentration"), name = "ΔC") +
  theme_void(base_size = 14) +
  theme(legend.position = "bottom", legend.key.width = unit(1.4, "cm"))
```
When precinct data is present, the next three figures repeat the same three views for the focal party
vote, where segment color and ΔC now measure the focal party's share of the two-party vote instead of
the group population, so the same district can read differently on race and on votes.

Figure \@ref(fig:scatterplot-cpc-votes) is the partisan scatter. Each district sits by the share of
its vote inside the overlap region on the horizontal axis against its partisan ΔC on the vertical
axis, so a district high on the axis concentrates the focal party's vote relative to its compact
center and one low on it dilutes it.

```{r scatterplot-cpc-votes, echo = FALSE, eval = have_votes, fig.width = 7, fig.height = 6, fig.cap = paste0("CPC overlap share versus partisan ΔC by district, ", params$state, ".")}
ggplot(districts_results, aes(x = cpcIdentityPctV, y = cpcDiffV)) +
  geom_hline(yintercept = 0, color = "grey70") +
  geom_point(aes(fill = ifelse(cpcDiffV >= 0, "Concentration", "Dilution")),
             shape = 21, size = 9, color = "grey30", stroke = 0.6) +
  geom_text(aes(label = DISTRICT), color = "white", fontface = "bold", size = 4) +
  scale_fill_manual(values = c("Concentration" = "#DC6D85", "Dilution" = "#6EA6C3"), name = NULL) +
  labs(x = "CPC overlap share of votes (%)", y = "ΔC (percentage points)") +
  theme_bw(base_size = 14) +
  theme(legend.position = "top")
```

Figure \@ref(fig:cpc-viz-votes) is the same dumbbell for the focal party vote, read the same way. The
line at 0 is the overlap region, the circle is the captured region and the diamond the excluded region,
each at the focal party's share relative to the overlap, and the row is red where the district
concentrates the focal party's vote and blue where it dilutes it. Marker area is the region's two-party
vote, so a large marker carries many votes.

```{r cpc-viz-votes, echo = FALSE, eval = have_votes, warning = FALSE, fig.height = max(5, 0.9 * n_districts), fig.width = 10, fig.cap = paste0("CPC decomposition for the ", fp_disp, " vote into overlap, captured, and excluded regions, ", params$state, ". Read together with the ΔC map that follows.")}
make_cpc_dumbbell("cpcidPctV", "cpccaPctV", "cpcexPctV",
                  "cpccatotV", "cpcextotV", "cpcDiffV",
                  paste0(fp_disp, " share"))
```

Figure \@ref(fig:cpc-viz-votes-map) is the locator for the partisan bars. It shades each district on
the state map by its partisan ΔC, red for concentration and blue for dilution, so you can see where
the focal party's vote is concentrated or diluted across the state.

```{r cpc-viz-votes-map, echo = FALSE, eval = have_votes, warning = FALSE, fig.width = map_w, fig.height = map_w * map_asp + 1.0, fig.cap = paste0("Partisan ΔC by district for the ", fp_disp, " vote, ", params$state, ". The locator for the CPC decomposition above.")}
cpc_map_df <- districts |> dplyr::select(DISTRICT) |>
  left_join(dplyr::select(st_drop_geometry(districts_results), DISTRICT, cpcDiffV),
            by = "DISTRICT")
cpc_map_lim <- max(abs(cpc_map_df$cpcDiffV), na.rm = TRUE)
ggplot(cpc_map_df) +
  geom_sf(aes(fill = cpcDiffV), color = "grey40", linewidth = 0.2) +
  geom_sf_text(aes(label = DISTRICT), size = 4.5, fontface = "bold", color = "grey15") +
  scale_fill_gradient2(low = "#6EA6C3", mid = "white", high = "#DC6D85", midpoint = 0,
                       limits = c(-cpc_map_lim, cpc_map_lim),
                       breaks = c(-cpc_map_lim, 0, cpc_map_lim),
                       labels = c("Dilution", "0", "Concentration"), name = "ΔC") +
  theme_void(base_size = 14) +
  theme(legend.position = "bottom", legend.key.width = unit(1.4, "cm"))
```


```{r save-districts}
# the geometry layer here, the flat CSV once in the save-results chunk below
districts_results |> st_write(file.path(output_dir, "district_results.gpkg"),
                              append = FALSE)
```


Figure \@ref(fig:map-cpc) draws the compact population-center districts over the actual districts, so
the reference shape behind the ΔC is visible on the map. The white dot marks each district's
population-weighted center and the cross marks its geometric centroid, and the gap between the two
shows how far the population is pulled to one side of the district.

```{r map-cpc, echo = FALSE, warning = FALSE, fig.width = map_w, fig.height = map_w * map_asp + 0.6, fig.cap = paste0("Compact population-center districts over the actual districts, ", params$state, ".")}
# population-weighted center (from centers, in mbc_crs) and geometric centroid, both to 4269
weighted_ctr <- st_as_sf(centers, coords = c("cx", "cy"), crs = mbc_crs) |>
  st_transform(4269)
geom_ctr <- st_centroid(st_geometry(districts))
st_write(weighted_ctr, file.path(output_dir, "weighted_centers.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)

ggplot() +
  geom_sf(data = districts, fill = NA, color = "grey60", linewidth = 0.3) +
  geom_sf(data = cpc_districts, aes(fill = factor(cpcDistrict)), color = NA, alpha = 0.5) +
  geom_sf(data = geom_ctr, shape = 3, color = "grey15", size = 2, stroke = 0.7) +
  geom_sf(data = weighted_ctr, shape = 21, fill = "white", color = "grey15", size = 2.4, stroke = 0.7) +
  labs(fill = "District",
       subtitle = "White dot is the population-weighted center, cross is the geometric centroid.") +
  theme_void() +
  theme(legend.position = "bottom", plot.margin = margin(0, 0, 0, 0))
```


# 5 Results

## 5.1 Composition difference (ΔC) table

We report the composition difference two ways, first for the focal party vote and then for the
group population, matching the order of the figures. `r if (have_votes) "Tables \\@ref(tab:results-table-party) and \\@ref(tab:results-table) hold" else "Table \\@ref(tab:results-table) holds"` the
numeric core of the analysis, and every figure below draws from these numbers. Each row is a
district. `pctMinoritybg` is the group's share of the actual district, and `chPct`, `mbcPct`, and
`cpcPct` are the group's share inside the convex hull, the minimum bounding circle, and the compact
population center. `chDiff`, `mbcDiff`, and `cpcDiff` are the actual share minus each of those, so a
positive value is concentration and a negative value is dilution. To read it, scan the three ΔC columns
for each district. A district with large positive values across all three is strongly packed, and
one with large negative values across all three is strongly cracked. Values near zero mean the shape
does not move the composition much. When the three columns disagree in sign, the result depends on
the reference shape, which is itself worth noting. When precinct data is loaded, the partisan table
repeats the same calculation for the focal party vote, with `pctMP` the party's share of the
actual district and `chDiffV`, `mbcDiffV`, and `cpcDiffV` the differences from each reference shape.
Shape compactness is a separate question, reported in Table \@ref(tab:compactness-table). The full
result, including every column, is written to a CSV file.

```{r results-table-party, echo = FALSE, eval = have_votes}
show_cols_v <- c("DISTRICT", "pctMP", "chPctV", "chDiffV", "mbcPctV", "mbcDiffV",
                 "cpcPctV", "cpcDiffV")
tabv <- districts_results |> st_drop_geometry() |>
  dplyr::select(any_of(show_cols_v)) |>
  arrange(DISTRICT)
save_output_table(tabv, paste0("Delta-C by district for the ", fp_disp, " vote"))
tabv |>
  kbl(caption = paste0("ΔC by district for the ", fp_disp, " vote, ",
                         params$state, "."),
      col.names = c("District", "Actual %", "Share %", "ΔC", "Share %", "ΔC", "Share %", "ΔC")) |>
  style_tbl() |>
  column_spec(4, bold = TRUE, color = diff_color(tabv$chDiffV)) |>
  column_spec(6, bold = TRUE, color = diff_color(tabv$mbcDiffV)) |>
  column_spec(8, bold = TRUE, color = diff_color(tabv$cpcDiffV)) |>
  add_header_above(ref_header, escape = FALSE) |>
  footnote(general = paste0("ΔC is the actual district's share minus the reference share. ",
             "Positive is <span style='color:#C0405E'>concentration</span> and negative ",
             "is <span style='color:#3E7EA0'>dilution</span>."),
           general_title = "Note. ", escape = FALSE) |>
  scroll_box(width = "100%")
```

```{r results-table, echo = FALSE}
show_cols <- c("DISTRICT", "pctMinoritybg", "chPct", "chDiff", "mbcPct", "mbcDiff",
               "cpcPct", "cpcDiff")
tab <- districts_results |> st_drop_geometry() |>
  dplyr::select(any_of(show_cols)) |>
  arrange(DISTRICT)
save_output_table(tab, paste0("Delta-C by district for the ", params$group, " ", basis_word))
tab |>
  kbl(caption = paste0("ΔC by district for the ", params$group, " ", basis_word, ", ",
                         params$state, "."),
      col.names = c("District", "Actual %", "Share %", "ΔC", "Share %", "ΔC", "Share %", "ΔC")) |>
  style_tbl() |>
  column_spec(4, bold = TRUE, color = diff_color(tab$chDiff)) |>
  column_spec(6, bold = TRUE, color = diff_color(tab$mbcDiff)) |>
  column_spec(8, bold = TRUE, color = diff_color(tab$cpcDiff)) |>
  add_header_above(ref_header, escape = FALSE) |>
  footnote(general = paste0("ΔC is the actual district's share minus the reference share. ",
             "Positive is <span style='color:#C0405E'>concentration</span> and negative ",
             "is <span style='color:#3E7EA0'>dilution</span>."),
           general_title = "Note. ", escape = FALSE) |>
  scroll_box(width = "100%")
```

```{r save-results}
districts_results |> st_drop_geometry() |>
  write_csv(file.path(output_dir, "district_results.csv"))
```

## 5.2 Interpreting the composition difference (ΔC)

Our approach asks a single question, whether a district's shape changes who it
represents. Rather than judging a district by how irregular it looks, it compares the
composition of the actual district to the composition of a compact reference version of
that same district. The reference can be the convex hull, the minimum bounding circle,
or the compact population center. The metric is the difference between the two
compositions.

$$\Delta C = C_{district} - C_{reference}$$

Here $C_{district}$ is the minority percentage inside the actual district and
$C_{reference}$ is the minority percentage inside the compact reference shape. A subscript
names the reference shape, so $\Delta C_{hull}$, $\Delta C_{mbc}$, and $\Delta C_{cpc}$ apply
this formula to the convex hull, the minimum bounding circle, and the compact population
center (in code `chDiff`, `mbcDiff`, and `cpcDiff`). When precinct data is present a
superscript $v$ marks the version computed on the focal party vote, so $\Delta C_{hull}^{v}$,
$\Delta C_{mbc}^{v}$, and $\Delta C_{cpc}^{v}$ (in code `chDiffV`, `mbcDiffV`, and `cpcDiffV`).
The sign carries the direction, a positive $\Delta C$ toward concentration and a negative one
toward dilution. We use concentration and dilution as neutral terms for the direction, and reserve
packing and cracking for districts with clear evidence of gerrymandering.

There are three cases to read. When $\Delta C$ is strongly positive, the actual district
holds a higher share of the group than its compact reference, which we call concentration.
This may indicate packing where the district is already majority-minority and the group is
over-represented relative to the compact shape. When $\Delta C$ is strongly negative, the
actual district holds a lower share than its compact reference, which we call dilution, and
which may indicate cracking or the exclusion of nearby group members. When $\Delta C$ is close to zero, the
irregular shape does not meaningfully change the group's composition, so the district may
look non-compact while its shape does not appear to distort racial or partisan
representation.

In the figures that follow, the reading is based on the distance between the actual
district (the black point) and the compact-shape estimates. The black point to the right
of the others points toward concentration, and to the left points toward dilution. As you read
the table and figures, consider which districts show the largest gap between the actual
composition and the compact-shape composition, which appear potentially concentrated, which
appear potentially diluted, and which stay similar across all geometries.

## 5.3 Composition by district and reference shape

Figure \@ref(fig:figure-composition) has one panel per district. Within a panel, the row labeled Minority Population (%) is the
racial or ethnic composition, and the row labeled Focal Party Vote (%) is the partisan
composition. Minority Population is the share of the `r basis_word` that belongs to the
group set by `params$group`. Focal Party Vote is the share of the two-party vote for the
chosen office won by the focal party. The vote row appears only when precinct data is
loaded.

Each row carries four points, one for the actual district and one for each compact reference
shape. Black dots are the composition of the actual district. Green dots are the composition
inside its convex hull. Purple dots are the composition inside its minimum bounding circle. Orange
dots are the composition inside its compact population center. The distance from the black dot to
each colored dot is that reference's ΔC.

To read it, go left to right along each row. When the black point sits to the right of the
colored points, the actual district holds more of the group than its compact references, which
points to concentration. When it sits to the left, the district holds less, which points to dilution.
When it sits on top of the colored points, the shape barely changes the composition. When
the colored points fall on both sides of the black point, the three methods disagree, so read that
district with care. The dashed lines mark margin bands around a majority, a tight one at 47.5 and 52.5 percent and a wider one at 45 and 55 percent, so a district inside them is close to even and only outside is the majority clear. We do not print the ΔC
numbers on the plot to keep it readable, so read the exact values from the ΔC table above.

```{r figure-composition, echo = FALSE, warning = FALSE, fig.height = max(6, 1.5 * n_districts), fig.width = 10, fig.cap = paste0("Composition by district and reference shape, ", params$state, ".")}
combined <- districts_results |> st_drop_geometry() |>
  dplyr::select(DISTRICT, actual = pctMinoritybg, hull = chPct, mbc = mbcPct, cpc = cpcPct) |>
  mutate(type = "Minority Population (%)")
if (have_votes)
  combined <- bind_rows(combined,
    districts_results |> st_drop_geometry() |>
      dplyr::select(DISTRICT, actual = pctMP, hull = chPctV, mbc = mbcPctV, cpc = cpcPctV) |>
      mutate(type = "Focal Party Vote (%)"))
combined <- combined |>
  mutate(DISTRICT = factor(DISTRICT, levels = sort(unique(districts_results$DISTRICT))),
         type = factor(type, levels = c("Minority Population (%)", "Focal Party Vote (%)")))

# long form so the reference type becomes a legend (color + shape), open shapes stay
# readable where points overlap
combined_long <- combined |>
  pivot_longer(c(actual, hull, mbc, cpc), names_to = "reference", values_to = "pct") |>
  mutate(reference = factor(reference, levels = c("actual", "hull", "mbc", "cpc"),
                            labels = c("Actual district", "Convex hull",
                                       "Min. bounding circle", "Compact pop. center")))

ggplot(combined_long, aes(x = pct, y = type, color = reference, shape = reference)) + xlim(0, 100) +
  geom_vline(xintercept = c(45, 55), color = "grey70", linewidth = 0.6, linetype = "dashed") +
  geom_vline(xintercept = c(47.5, 52.5), color = "grey35", linewidth = 0.7, linetype = "dashed") +
  geom_point(size = 3.8, stroke = 1.3) +
  scale_color_manual(values = c("Actual district" = "#222222", "Convex hull" = "#3E9B63",
                                "Min. bounding circle" = "#7D3C98", "Compact pop. center" = "#D98400"),
                     name = NULL) +
  scale_shape_manual(values = c("Actual district" = 1, "Convex hull" = 2,
                                "Min. bounding circle" = 0, "Compact pop. center" = 5),
                     name = NULL) +
  facet_wrap(~ DISTRICT, ncol = 1, labeller = label_both) +
  labs(x = "Percentage (%)", y = NULL,
       caption = "Dashed grey lines mark margin bands around a 50% majority, tight at 47.5 to 52.5 and wide at 45 to 55.") +
  theme_bw(base_size = 15) +
  theme(legend.position = "top", plot.caption = element_text(hjust = 0, color = "grey35"))
```

## 5.4 Composition difference (ΔC) by district

We show the ΔC two ways, first as bars for the exact size and ranking, then as maps for the
spatial pattern.

### 5.4.1 ΔC bar chart by district and reference shape

Each bar in `r if (have_votes) "Figures \\@ref(fig:figure-cdiff-bars-party) and \\@ref(fig:figure-cdiff-bars-race)" else "Figure \\@ref(fig:figure-cdiff-bars-race)"` is one district's ΔC. Bars to the right are concentration and bars to the left are
dilution. The three panels are the three reference shapes, in the same order and colors as the
composition figure above. Districts are sorted by convex hull ΔC, so the most packed sit at
the top and the most cracked at the bottom. `r if (have_votes) "The partisan version comes first and the racial version below it." else "Only the racial version is shown, because no precinct data is loaded."` A packed
district shows a right-pointing bar in every panel, and a cracked district shows a left-pointing
bar in every panel. When a district's bar points right in one panel but left in another, the
reference shapes disagree, which usually means the population is naturally clustered there, so
treat the case with care.

```{r cdiff-bars-setup, include = FALSE}
ord <- districts_results |> st_drop_geometry() |> arrange(chDiff) |> pull(DISTRICT)

make_cdiff_bars <- function(dcols, ttl) {
  tab <- districts_results |> st_drop_geometry() |> dplyr::select(DISTRICT, dplyr::all_of(dcols))
  names(tab)[2:4] <- c("Convex Hull", "Min. Bounding Circle", "Compact Pop. Center")
  tab |>
    pivot_longer(-DISTRICT, names_to = "reference", values_to = "Cdiff") |>
    mutate(DISTRICT = factor(DISTRICT, levels = ord),
           reference = factor(reference, levels = c("Convex Hull", "Min. Bounding Circle", "Compact Pop. Center")),
           direction = ifelse(Cdiff >= 0, "Concentration", "Dilution")) |>
    ggplot(aes(x = Cdiff, y = DISTRICT, fill = direction)) +
    geom_col() +
    geom_vline(xintercept = 0, color = "grey35", linewidth = 0.6) +
    scale_fill_manual(values = c("Concentration" = "#DC6D85", "Dilution" = "#6EA6C3")) +
    facet_wrap(~ reference, labeller = as_labeller(ref_labels)) +
    labs(x = "ΔC (percentage points)", y = "District", fill = NULL) +
    theme_bw(base_size = 15) +
    theme(legend.position = "bottom",
          strip.text = ggtext::element_markdown(size = 15, face = "bold"))
}
```

```{r figure-cdiff-bars-party, echo = FALSE, eval = have_votes, fig.height = max(5, 0.9 * n_districts), fig.width = 11, fig.cap = paste0("ΔC by district for the ", fp_disp, " vote, by reference shape, ", params$state, ".")}
make_cdiff_bars(c("chDiffV", "mbcDiffV", "cpcDiffV"),
                paste0("ΔC by district, focal party vote (", params$state, ")"))
```

```{r figure-cdiff-bars-race, echo = FALSE, fig.height = max(5, 0.9 * n_districts), fig.width = 11, fig.cap = paste0("ΔC by district for the ", params$group, " ", basis_word, ", by reference shape, ", params$state, ".")}
make_cdiff_bars(c("chDiff", "mbcDiff", "cpcDiff"),
                paste0("ΔC by district, ", params$group, " population (", params$state, ")"))
```

### 5.4.2 ΔC choropleth map by district

`r if (have_votes) "Figures \\@ref(fig:figure-cdiff-map-party) and \\@ref(fig:figure-cdiff-map-race)" else "Figure \\@ref(fig:figure-cdiff-map-race)"` draw the same ΔC on the district map, so the spatial pattern is visible. Each district is
shaded by its ΔC, red for concentration and blue for dilution, with each district number labeled. The
three panels are the three reference shapes, side by side, in the same order as the bar chart. `r if (have_votes) "As with the bars, the partisan map comes first and the racial map below it." else "Only the racial map is shown, because no precinct data is loaded."` To read it, find a district by its number and compare
its color across the three panels. A district that is red in all three panels is robustly concentrated,
and one that is blue in all three is robustly diluted. A district that changes color between
panels is a case where the choice of reference shape matters.

```{r cdiff-map-setup, include = FALSE}
make_cdiff_map <- function(dcols, ttl) {
  tab <- districts_results |> st_drop_geometry() |> dplyr::select(DISTRICT, dplyr::all_of(dcols))
  names(tab)[2:4] <- c("Convex Hull", "Min. Bounding Circle", "Compact Pop. Center")
  tab <- tab |>
    pivot_longer(-DISTRICT, names_to = "reference", values_to = "Cdiff") |>
    mutate(reference = factor(reference, levels = c("Convex Hull", "Min. Bounding Circle", "Compact Pop. Center")))
  m <- districts |> dplyr::select(DISTRICT) |> left_join(tab, by = "DISTRICT")
  lim <- max(abs(m$Cdiff), na.rm = TRUE)   # symmetric color range
  ggplot(m) +
    geom_sf(aes(fill = Cdiff), color = "grey30", linewidth = 0.2) +
    # district number on top, signed ΔC below it: a positive (red) value is
    # concentration and a negative (blue) value is dilution, read per district and panel
    geom_sf_text(aes(label = paste0(DISTRICT, "\n",
                     ifelse(is.na(Cdiff), "", sprintf("%+.1f", Cdiff)))),
                 size = 3.8, fontface = "bold", color = "grey10", lineheight = 0.8) +
    scale_fill_gradient2(low = "#6EA6C3", mid = "white", high = "#DC6D85",
                         midpoint = 0, limits = c(-lim, lim),
                         breaks  = c(-lim, 0, lim),
                         labels  = c("Dilution", "0", "Concentration"),
                         name = "ΔC") +
    facet_wrap(~ reference, ncol = 3, labeller = as_labeller(ref_labels)) +
    # horizontal colorbar along the bottom, so the "Dilution"/"Concentration" end
    # labels are not clipped in wide, short states
    guides(fill = guide_colorbar(title.position = "left", title.vjust = 0.9,
                                 barwidth = unit(9, "cm"), barheight = unit(0.5, "cm"))) +
    theme_void(base_size = 18) +
    theme(strip.text = ggtext::element_markdown(size = 20, face = "bold", margin = margin(b = 6)),
          legend.position = "bottom",
          legend.title = element_text(size = 18),
          legend.text = element_text(size = 16))
}
```

```{r figure-cdiff-map-party, echo = FALSE, eval = have_votes, fig.width = 12, fig.height = 4 * map_asp + 2.0, fig.cap = paste0("ΔC mapped for the ", fp_disp, " vote, by reference shape, ", params$state, ".")}
make_cdiff_map(c("chDiffV", "mbcDiffV", "cpcDiffV"),
               paste0("ΔC by district, focal party vote (", params$state, ")"))
```

```{r figure-cdiff-map-race, echo = FALSE, fig.width = 12, fig.height = 4 * map_asp + 2.0, fig.cap = paste0("ΔC mapped for the ", params$group, " ", basis_word, ", by reference shape, ", params$state, ".")}
make_cdiff_map(c("chDiff", "mbcDiff", "cpcDiff"),
               paste0("ΔC by district, ", params$group, " population (", params$state, ")"))
```
